---
title: "PBR Interpretation based on Frick et al. 2026 (`r params$project_ref`)"
output:
  word_document:
    toc: true
params:
  project_ref: ""
  species_name: ""
  species_common_name: ""
  thresholds: !r NULL
  fr_original: !r NULL
  fr_corrected: !r NULL
  iucn_status_original_guess: ""
  iucn_status_corrected: ""
  backsolve_grid: !r NULL
  nmin_assumed: !r NULL
  project_footprint_km2: !r NULL
  combined_region_km2: !r NULL
  interproject_distance_km: !r NULL
  nmin_plausibility: !r NULL
  density_proxy_low: !r NULL
  density_proxy_high: !r NULL
  pbr_corrected_fr: !r NULL
  s_range: !r NULL
  alpha_range: !r NULL
  alpha_range_prior: !r NULL
  elasticity_ranges: !r NULL
  fig_lambda_max: ""
  fig_elasticity_survival: ""
  fig_elasticity_alpha: ""
  fig_response_surfaces: ""
  fig_response_surfaces_3d: !r NULL
  fig_response_surfaces_3d_html: !r NULL
  fig_pbr_density: ""
  n_sim: !r NULL
  pbr_quantiles: !r NULL
  threshold_percentiles: !r NULL
  alpha_range_comparison: !r NULL
  leslie_s_juv: !r NULL
  leslie_s_adult_f: !r NULL
  leslie_s_adult_m: !r NULL
  leslie_p_breed: !r NULL
  leslie_litter: !r NULL
  leslie_F: !r NULL
  lambda_leslie_validated: !r NULL
  leslie_vs_pbr: !r NULL
  fig_leslie_maturation: ""
  n0_total: !r NULL
  pva_n_years: !r NULL
  pva_n_reps: !r NULL
  pva_vital_rate_cv: !r NULL
  quasi_ext_threshold: !r NULL
  pva_risk_summary: !r NULL
  fig_pva_projection: ""
  comparison_summary: !r NULL
  leslie_breakeven_tbl: !r NULL
  leslie_boundary_band_shares: !r NULL
  leslie_boundary_pct_below_baseline: !r NULL
  leslie_boundary_range_s_adult: !r NULL
  leslie_boundary_range_s_juv: !r NULL
  leslie_boundary_range_p_breed: !r NULL
  leslie_boundary_range_litter: !r NULL
  fig_leslie_boundary_surface: ""
  nmin_vs_lambda_sensitivity: !r NULL
  fig_nmin_lambda_surface: ""
  real_correction_factor: !r NULL
  real_curtailment_start: ""
  real_curtailment_week: !r NULL
  real_checkpoint_week: !r NULL
  real_overage_summary: !r NULL
  real_yoy_summary: !r NULL
  real_forecast_summary: !r NULL
  real_n_turbines_by_project: !r NULL
  real_turbine_hist: !r NULL
  fig_real_dashboard: ""
  fig_real_turbine_pareto: ""
  turbine_validation_correction_ratio_range: !r NULL
  turbine_validation_correction_ratio_median: !r NULL
  fig_turbine_validation: ""
  turbine_validation_bash_window: ""
  turbine_validation_dzh_window: ""
  turbine_validation_bash_data_start: ""
  turbine_validation_dzh_data_start: ""
  turbine_validation_comparison: !r NULL
  turbine_validation_overlap: !r NULL
  turbine_validation_correction_check: !r NULL
  curtailment_year_data_max_date: ""
  curtailment_year_checkpoint_week: !r NULL
  curtailment_year_reduction_so_far: !r NULL
  curtailment_year_full_counterfactual: !r NULL
  fig_curtailment_year: ""
  global_nmin_area_km2: !r NULL
  global_density_low: !r NULL
  global_density_high: !r NULL
  global_pbr_quantiles_by_project: !r NULL
  global_pbr_old_new_comparison: !r NULL
  fig_global_pbr_uncertainty: ""
  candidate_pbr_table: !r NULL
  candidate_pbr_effectiveness: !r NULL
  fig_candidate_pbr: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
# knitr::include_graphics() defaults to rewriting absolute paths as
# relative to rmarkdown::render()'s output_dir (here "outputs/", not this
# .Rmd's own "report/" folder) -- pandoc then can't resolve the result and
# silently swaps in the image's alt-text description instead of embedding
# it. The figure paths passed in via params are already absolute; keep
# them as-is.
options(knitr.graphics.rel_path = FALSE)
```

Document history / List of changes.

| Version | Changes | Date of changes | Authors |
|---|---|---|---|
| Version 1 | First draft | 2026-05-07 | Paulo Cardoso |
| Version 2 | Corrected recovery factor (Fr); reverse-engineered and plausibility-checked the local population estimate (Nmin); updated age-at-first-breeding range from a literature review; rebuilt the sensitivity, response-surface and Monte Carlo analyses on consistent, reproducible parameter ranges | 2026-09 | Paulo Cardoso |
| Version 3 | Added a Leslie-matrix cross-check (validated against the published Dekker & Limpens 2024 model) and a 25-year stochastic PVA-lite projection, with a summary comparison against the PBR/Niel-Lebreton approach | 2026-09 | Paulo Cardoso |
| Version 4 | Added an inverse demographic boundary analysis on the validated Leslie matrix, and a discussion of Nmin as a source of uncertainty, including Frick et al.'s own definition of local population size and its stated limitation for cumulative impacts across nearby facilities | 2026-09 | Paulo Cardoso |
| Version 5 | Replaced the kaleido-based 3D response-surface export with a fixed-camera static image; anonymised facility names to Project 1/Project 2 for external presentation; added a direct quantitative comparison of PBR's sensitivity to Nmin versus lambda_max, and a synthesising Nmin x lambda_max surface figure | 2026-09 | Paulo Cardoso |
| Version 6 | Restructured section hierarchy (Leslie matrix cross-check and the closing interpretation promoted to top-level sections; PVA-lite, the three-way comparison, the boundary analysis and the Nmin discussion grouped under a new "Comparing approaches" section); improved legibility of four figures; rewrote the closing section as "Integrated interpretation and implications", synthesising the full body of evidence rather than the PBR/Niel-Lebreton sensitivity analysis alone | 2026-09 | Paulo Cardoso |
| Version 7 | Promoted PVA-lite to its own top-level section (4), ahead of "Comparing approaches" (5, now containing the three-way comparison, the boundary analysis, the Nmin discussion and the capacity-vs-population-size synthesis as 5.1-5.4); "Integrated interpretation and implications" is now section 6 | 2026-09 | Paulo Cardoso |
| Version 8 | Added section 7, "Towards operational adaptive mortality management" -- a proposed methodology for converting the annual PBR reference into a weekly PCFM-driven control chart and turbine-level curtailment allocation, illustrated on synthetic data pending the real weekly PCFM dataset | 2026-09 | Paulo Cardoso |
| Version 9 | Section 7 now applies the adaptive-management mechanism to the real 2025-2026 PCFM carcass data (V. murinus, GenEst-corrected), in place of the earlier synthetic illustration: year-to-date mortality vs. the annual reference, a year-over-year comparison, a forward forecast, and pre-curtailment turbine-level concentration (turbine identifiers anonymised) | 2026-09 | Paulo Cardoso |
| Version 10 | Corrected the imposed annual thresholds throughout from 144/120 to 112/88 bats/year, the values actually used operationally, confirmed against the Blanket Curtailment Plan technical note | 2026-09 | Paulo Cardoso |
| Version 11 | Added two turbine- and season-level checks against real operational data: a retrospective validation of an independently-derived turbine selection against the actual, lender-agreed selection documented in the Blanket Curtailment Plan; and a full operational-year (May 2026-May 2027) comparison of actual, post-curtailment mortality against a no-curtailment counterfactual built from the prior year's pattern | 2026-09 | Paulo Cardoso |
| Version 12 | Fixed a future-dated PCFM record (data-entry typo) that had pulled Section 7.4's checkpoint 2-3 weeks ahead of the real data; clarified why Section 7.2's turbine universe differs from the official selection's; added Section 7.5, "2027 mitigation frontier": extends the PBR Monte Carlo with project-specific Nmin uncertainty (in place of the fixed reverse-engineered value used elsewhere), then compares the current thresholds against Q05/Q25/median/Q75 of the resulting distribution on identical terms -- PVA-lite risk, mortality-budgeting reduction required, and turbines needed under a literature-benchmarked curtailment-effectiveness assumption (Adams, Gulka & Williams 2021) -- as decision-support for the 2027 season, not a proposal to change the imposed thresholds | 2026-09 | Paulo Cardoso |

# Context

This technical note reviews the Potential Biological Removal (PBR)
framework proposed by Frick et al. (2026) for bats at wind-energy
facilities, and applies it to the `r params$species_common_name` (*`r params$species_name`*)
fatality thresholds currently imposed at `r params$project_ref`
(`r paste(sprintf("%s bats/year at %s", params$thresholds$threshold, params$thresholds$facility), collapse = "; ")`).
No documented derivation of these threshold values is available, and no
independent local population estimate for the species in this region
exists or can realistically be obtained. The note therefore works
backwards from the imposed thresholds to identify a parsimonious,
checkable set of assumptions consistent with them, and then tests how
sensitive those thresholds are to reasonable demographic uncertainty,
rather than to the single fixed parameter values the framework uses by
default.

# PBR Threshold Uncertainty

## Frick et al. adjusted PBR framework

The PBR framework presented by Frick and colleagues (Frick et al. 2026,
herein referred to as Frick) adapts the classical Potential Biological
Removal concept to bat mortality at wind-energy facilities. It retains the
precautionary structure of the original PBR formulation while addressing
known limitations of applying marine-mammal-derived PBR methods to volant
terrestrial taxa, particularly long-lived bat species characterised by low
fecundity, delayed maturity, and strong sensitivity to adult survival.

Under the adjusted framework, PBR is expressed as:

$$PBR = \frac{1}{2} R_{max} \cdot F_r \cdot N_{min}$$

where $R_{max} = \lambda_{max} - 1$ is the maximum annual population growth
rate, $F_r$ is a recovery factor linked to IUCN conservation status, and
$N_{min}$ is a conservative lower-bound estimate of local abundance.

A key conceptual shift in the Frick proposal is the move from broad
regional or continental population framing towards a project-scale local
population proxy. Local abundance bounds ($N_L$ and $N_U$) are derived from
the interaction area surrounding the wind facility combined with plausible
density estimates, and a lognormal formulation is then used to derive the
geometric mean abundance estimate ($\hat{N}$), its coefficient of variation
($CV_{\hat{N}}$), and ultimately $N_{min}$. Uncertainty is therefore
incorporated explicitly into the abundance term: poorly constrained
abundance estimates are penalised with lower, more precautionary PBR
values.

$\lambda_{max}$ is standardised using the Niel & Lebreton (2005)
demographic-invariant approximation, from adult survival ($s$) and age at
first breeding ($\alpha$):

$$\lambda_{max} \approx \frac{(s\alpha - s + \alpha + 1) + \sqrt{(s - s\alpha - \alpha - 1)^2 - 4s\alpha^2}}{2\alpha}$$

Frick et al. propose precautionary fixed $\lambda_{max}$ values (1.24 for
most bats, 1.20 for families with later age at first breeding such as
Pteropodidae) to reduce sensitivity to poorly constrained demographic
inputs. The framework remains structurally dependent on several uncertain
ecological assumptions, including the definition of the local interacting
population, density estimation, recovery-factor selection, and the
demographic assumptions underlying the fixed $\lambda_{max}$ values.
Accordingly, PBR thresholds are best interpreted as precautionary decision
bands rather than deterministic operational limits, particularly where
they are used to calibrate mitigation measures such as curtailment.

## Sensitivity and elasticity analysis

Sensitivity and elasticity analyses were undertaken to evaluate how
uncertainty in the two demographic parameters used by the Niel & Lebreton
approximation, adult survival ($s$) and age at first breeding ($\alpha$),
propagates into $\lambda_{max}$ and consequently into PBR. Parameter ranges
were set at $s \in$ [`r params$s_range[1]`, `r params$s_range[2]`] and
$\alpha \in$ [`r params$alpha_range[1]`, `r params$alpha_range[2]`] years,
the latter following the literature review below. For each combination,
$\lambda_{max}$ was recalculated and local sensitivity estimated
numerically using finite differences. Elasticity is a dimensionless
proportional sensitivity metric: the expected percentage variation in
$\lambda_{max}$ resulting from a 1% variation in a demographic parameter.

```{r fig-lambda-max}
knitr::include_graphics(params$fig_lambda_max)
```

Within the explored range, $\lambda_{max}$ *decreases* as adult survival or
age at first breeding increase -- a counterintuitive relationship
documented by Niel & Lebreton (2005) and noted by Frick et al. themselves,
arising from the life-history trade-off between survival and fecundity
built into the demographic-invariant approximation. This is a decrease in
absolute terms, not a sign change: for any survival probability in (0, 1),
this formula's $\lambda_{max}$ only approaches 1 (a stable population) in
the limit as age at first breeding tends to infinity, and never actually
reaches it for a realistic age at first breeding -- the population is
growing ($\lambda_{max} > 1$) throughout the entire range explored here and
well beyond it. The elasticity values below should be read with that in
mind: a negative elasticity means $\lambda_{max}$ falls (while remaining
well above 1) as the parameter increases, not that $\lambda_{max}$ itself
becomes small or negative.

```{r elasticity-table}
knitr::kable(params$elasticity_ranges, digits = 3,
             caption = "Elasticity of lambda_max to each demographic parameter.")
```

```{r elasticity-inline}
es <- params$elasticity_ranges[params$elasticity_ranges$Parameter == "Adult survival (s)", ]
ea <- params$elasticity_ranges[params$elasticity_ranges$Parameter == "Age at first breeding (alpha)", ]
```

$\lambda_{max}$ is substantially more sensitive to uncertainty in adult
survival than to uncertainty in age at first breeding across the explored
demographic space: a 1% variation in adult survival produces roughly
`r round(abs(es$min), 2)`-`r round(abs(es$max), 2)`% variation in
$\lambda_{max}$, versus `r round(abs(ea$min), 2)`-`r round(abs(ea$max), 2)`%
for an equivalent variation in age at first breeding. Sensitivity is not
spatially uniform: combinations of relatively high adult survival and
earlier reproductive maturity produce the largest elastic responses, which
is particularly relevant because adult survival is also among the least
well-constrained demographic parameters for most bat species.

```{r fig-elasticity-survival}
knitr::include_graphics(params$fig_elasticity_survival)
```

```{r fig-elasticity-alpha}
knitr::include_graphics(params$fig_elasticity_alpha)
```

These results support the central rationale of the Frick adjusted
framework: uncertainty in demographic parameterisation can materially
influence resulting PBR values, and precautionary fixed-$\lambda_{max}$
assumptions reduce excessive sensitivity to uncertain demographic inputs.
However, substantial uncertainty remains even under a fixed-$\lambda_{max}$
approach, through the interaction of demographic assumptions, local
abundance estimation, and recovery-factor selection -- the remainder of
this note works through that interaction for the `r params$species_common_name`
at `r params$project_ref` specifically.

## Evidence base for age at first breeding

For PBR, $\alpha$ should represent effective age at first reproduction --
the age at which a female actually recruits offspring into the population
-- not merely physiological sexual maturity. A literature review for
*`r params$species_name`* was undertaken on this basis.

| Source | Species / group | Relevant information | Implication for alpha |
|---|---|---|---|
| AnAge database | *Vespertilio murinus* | Female and male sexual maturity listed at 730 days (~2 years); litter size 2; one litter per year. | Supports alpha ~= 2 as a defensible value. |
| Haensel 2010, *Nyctalus* | *Vespertilio murinus* | Some females raise young in their first year of life; the author's own data suggest most or all females may reach sexual maturity in the year of birth and raise young as one-year-olds, stated cautiously. | Suggests alpha = 1-2 may be plausible; alpha = 2 is not an optimistic assumption. |
| Ratpenats species account | *Vespertilio murinus* | Many females mate in their first autumn. | Supports early sexual maturity, but mating is not the same as confirmed breeding. |
| Racey & Entwistle synthesis, cited in Frick et al.'s own discussion | Bats broadly | The general view for most bats is that age at first breeding is around 1 year, though first-year breeding probability can be lower; Frick et al. use alpha = 1.5 as a general benchmark and suggest alpha = 2 for some families. | Supports alpha = 1.5-2 as plausible for many bat species. |
| Cryan et al. 2012 | Migratory tree bats killed at wind turbines | Ovarian development indicating sexual maturity even in first-year females for several migratory tree-roosting species. | Useful analogue: some wind-vulnerable migratory bats mature very early. |
| Komar et al. 2020 | Male *Vespertilio murinus* | Food restriction delayed seasonal sexual maturation in males; well-fed males matured earlier. | Maturation timing is condition-dependent; male data are less directly useful for PBR (which uses female recruitment). |

Table: Sources reviewed for age at first breeding in the Parti-coloured
bat.

$\alpha = 2$ is a defensible, reasonably conservative central value for PBR
purposes under the effective-recruitment reading of $\alpha$, and is
directly supported by the AnAge sexual-maturity estimate. $\alpha = 1.5$
(the paper's own general bat benchmark) is a plausible lower bound, though
it rests on thinner, more cautiously stated species-specific evidence.
$\alpha = 3-4$ has no direct empirical support for *V. murinus* in the
sources reviewed; it remains a useful precautionary sensitivity scenario
for ecological constraints on effective recruitment (migration costs, poor
juvenile survival, low first-year breeding probability), but is not the
best-supported biological estimate. This review supports using
$\alpha \in$ [`r params$alpha_range[1]`, `r params$alpha_range[2]`] years as
the working range throughout this note, in place of a single fixed value.

## Reverse-engineering the imposed thresholds

`r params$project_ref` `r if (nrow(params$thresholds) > 1) "were" else "was"` given
a fixed annual fatality threshold for the `r params$species_common_name` --
`r paste(sprintf("%s bats/year at %s", params$thresholds$threshold, params$thresholds$facility), collapse = " and ")`
-- without a documented derivation. Since no independent local population
estimate for the species exists for this region, plausibility is assessed
by working backwards from the imposed thresholds under the Frick PBR
equation.

```{r backsolve-table}
knitr::kable(params$backsolve_grid, digits = c(0, 0, 2, 0),
             col.names = c("Facility", "Threshold (bats/yr)", "Fixed lambda_max", "Required Nmin"),
             caption = paste0("Local population size (Nmin) required to reproduce each imposed threshold, under Fr = ",
                               params$fr_original, " (", params$iucn_status_original_guess,
                               ") and the paper's fixed lambda_max benchmarks."))
```

Under $F_r=$ `r params$fr_original` (`r params$iucn_status_original_guess`)
and the paper's fixed $\lambda_{max}$ benchmarks applied per facility, the
four threshold x benchmark combinations above back-solve to Nmin values
ranging from ~`r format(round(min(params$backsolve_grid$Nmin_required)), big.mark = ",")`
to ~`r format(round(max(params$backsolve_grid$Nmin_required)), big.mark = ",")`
individuals -- no single population size reproduces both facilities'
thresholds under either fixed $\lambda_{max}$ benchmark. In the absence of
a shared or unique value, this note uses their median
(~`r format(params$nmin_assumed, big.mark = ",")`) as a single
representative working value for the density-plausibility check that
follows. This is a convenience for illustrative comparison, not evidence
that a population of that specific size was itself assumed when the
thresholds were originally derived: the parameters actually used for that
derivation are not documented and were not provided along with the
thresholds, and the absence of a shared value across the four
combinations is itself consistent with that -- if anything, it reinforces
that $N_{min}$ cannot be pinned down from the thresholds alone. `r params$project_ref`
are approximately `r params$interproject_distance_km` km apart, so if a
local population of this size is assumed at all, it is better read as two
similarly sized local populations rather than a single population shared
between them (see below).

Whether ~`r format(params$nmin_assumed, big.mark = ",")` individuals is a
plausible local population can be assessed against the real, known project
geography, by asking what average bat density it would imply, rather than
backing out a hypothetical area from a hypothetical density.

```{r nmin-plausibility-table}
knitr::kable(
  params$nmin_plausibility, digits = c(0, 0, 2, 2),
  col.names = c("Area scenario", "Area (km2)", "Implied density low (bats/km2)", "Implied density high (bats/km2)"),
  caption = paste0("Density that Nmin = ", format(params$nmin_assumed, big.mark = ","),
                    " would imply, under the two real area readings, holding the density ratio fixed at the default proxy's (",
                    params$density_proxy_low, "-", params$density_proxy_high, " bats/km2).")
)
```

The single-project-footprint reading
(`r params$nmin_plausibility$area_km2[1]` km$^2$) implies a density of
`r round(params$nmin_plausibility$density_low[1], 1)`-`r round(params$nmin_plausibility$density_high[1], 1)`
bats/km$^2$ -- closely matching the paper's own default density proxy
(`r params$density_proxy_low`-`r params$density_proxy_high` bats/km$^2$,
based on *Pipistrellus pipistrellus*, used in the absence of
species-specific density data) for a common bat species. Treating
$N_{min} \approx `r format(params$nmin_assumed, big.mark = ",")`$ as the
local population near each facility individually, rather than shared
across both, is therefore consistent with typical bat density and with
Frick et al.'s own project-scale definition of local population.

By contrast, treating
$N_{min} = `r format(params$nmin_assumed, big.mark = ",")`$ as a single
population spread across the combined region spanning both projects
(`r params$nmin_plausibility$area_km2[2]` km$^2$) implies a density of only
`r round(params$nmin_plausibility$density_low[2], 2)`-`r round(params$nmin_plausibility$density_high[2], 2)`
bats/km$^2$ -- roughly an order of magnitude below the default proxy, an
assumption that would need independent supporting evidence (e.g. that
regional density for `r params$species_common_name` here is genuinely that
sparse) before being considered reasonable.

The project-footprint reading is therefore both the more defensible order
of magnitude and the one that matches Frick et al.'s own project-scale
local-population concept. The working assumption throughout the rest of
this note is accordingly that $N_{min} \approx `r format(params$nmin_assumed, big.mark = ",")`$
applies to each facility's own local population, not to a single population
shared across the `r params$interproject_distance_km` km separating
`r paste(params$thresholds$facility, collapse = " and ")`.

$F_r = `r params$fr_original`$ corresponds to a `r params$iucn_status_original_guess`
status. The `r params$species_common_name` is not a threatened category, so
the defensible recovery factor is $F_r = `r params$fr_corrected`$
(`r params$iucn_status_corrected`). Holding $N_{min}$ at
`r format(params$nmin_assumed, big.mark = ",")` and correcting only $F_r$:

```{r pbr-corrected-fr-table}
knitr::kable(params$pbr_corrected_fr, digits = c(2, 0),
             col.names = c("Fixed lambda_max", "PBR (bats/yr)"),
             caption = paste0("PBR at Nmin = ", format(params$nmin_assumed, big.mark = ","),
                               ", Fr = ", params$fr_corrected, "."))
```

The recovery-factor correction alone shifts PBR upward at every fixed
$\lambda_{max}$ benchmark, before any change to $\lambda_{max}$ itself is
considered.

## PBR response surfaces

```{r fig-response-surfaces}
knitr::include_graphics(params$fig_response_surfaces)
```

```{r fig-response-surfaces-3d}
if (!is.null(params$fig_response_surfaces_3d)) {
  knitr::include_graphics(params$fig_response_surfaces_3d)
}
```

`r if (!is.null(params$fig_response_surfaces_3d_html)) paste0("A live, rotatable version of this figure (hover for exact values, toggle Fr scenarios) is provided as a standalone file, *", basename(params$fig_response_surfaces_3d_html), "*, alongside this document -- Word cannot embed an interactive plot directly, but the file opens in any browser without needing a server.")`

Across the plausible demographic range, the imposed-threshold contours
only appear for a narrow combination of relatively high adult survival and
low age at first breeding -- outside that narrow band, PBR departs from
the imposed thresholds even before changing $F_r$.

## Monte Carlo simulation outputs

Uncertainty in adult survival and age at first breeding was propagated
into $\lambda_{max}$ and PBR by simulation, holding
$N_{min} = `r format(params$nmin_assumed, big.mark = ",")`$ and
$F_r = `r params$fr_corrected`$ fixed, with
$s \sim U(`r params$s_range[1]`, `r params$s_range[2]`)$ and
$\alpha \sim U(`r params$alpha_range[1]`, `r params$alpha_range[2]`)$
(`r format(params$n_sim, big.mark = ",")` draws).

```{r mc-quantile-table}
q_tbl <- tibble::tibble(
  Quantile = names(params$pbr_quantiles),
  `PBR (bats/yr)` = round(params$pbr_quantiles, 1)
)
knitr::kable(q_tbl, caption = "Simulated PBR distribution quantiles.")
```

```{r mc-percentile-table}
knitr::kable(
  params$threshold_percentiles, digits = 2,
  col.names = c("Facility", "Threshold (bats/yr)", "Percentile rank in simulated PBR", "Median ratio", "P95 ratio"),
  caption = "Where the imposed thresholds fall within the simulated PBR distribution."
)
```

The imposed thresholds sit on the low, conservative side of the simulated
distribution, not the permissive side: median simulated PBR is
`r round(min(params$threshold_percentiles$median_ratio), 2)`-`r round(max(params$threshold_percentiles$median_ratio), 2)`x
the imposed thresholds, reaching
`r round(min(params$threshold_percentiles$p95_ratio), 2)`-`r round(max(params$threshold_percentiles$p95_ratio), 2)`x
at the 95th percentile.

```{r mc-alpha-comparison-table}
alpha_comp_names <- c(
  "Alpha range", "Q05", "Median", "Q95",
  sprintf("%% below %s (%s)", params$thresholds$threshold, params$thresholds$facility)
)
knitr::kable(
  params$alpha_range_comparison, digits = 1, col.names = alpha_comp_names,
  caption = paste0("Effect of the alpha range on the simulated PBR distribution (Fr = ",
                    params$fr_corrected, ", Nmin = ", format(params$nmin_assumed, big.mark = ","), ").")
)
```

Moving from the wider, less literature-grounded alpha range
(`r params$alpha_range_prior[1]`-`r params$alpha_range_prior[2]`) to the
range the evidence review supports
(`r params$alpha_range[1]`-`r params$alpha_range[2]`) shifts the simulated
PBR distribution further up, not down: the literature review therefore
reinforces, rather than weakens, the conclusion below.

```{r fig-pbr-density}
knitr::include_graphics(params$fig_pbr_density)
```

# Leslie matrix cross-check

The PBR/Niel-Lebreton approach is one specific method among several used to
estimate a population's sustainable growth rate: simpler methods (e.g. a
Leslie/Lefkovitch matrix) and more complex ones (full population viability
analysis, PVA) are also standard practice, and checking whether they agree
is itself informative. Unlike Niel & Lebreton's demographic-invariant
approximation, which is deliberately built to need only adult survival and
age at first breeding, a Leslie matrix requires an explicit juvenile
survival rate and fecundity -- inputs the PBR calculation above does not
use at all.

Species-specific values for the Parti-coloured bat were located: Safi
(2006) reports juvenile survival = `r params$leslie_s_juv`, adult female
survival = `r params$leslie_s_adult_f`, adult male survival =
`r params$leslie_s_adult_m`, and `r params$leslie_p_breed * 100`% of adult
females breeding; Zhigalin & Moskvitina (2017) report a litter size of
`r params$leslie_litter` at suburban colonies (2.7-2.9 at urban colonies).
Safi's survival figures are simple return rates (animals marked in year
*t* recaptured in *t+1*), not estimates from a modern open-population
mark-recapture model, and no confidence intervals were reported --
treated here as central working values, not precisely known rates.

This parametrisation was cross-checked against Dekker & Limpens (2024),
"Population dynamics of Dutch bats relevant to wind energy", which
independently built and published a *V. murinus* stage matrix from the
same Safi (2006) and Zhigalin & Moskvitina data. The female recruitment
term they use, $F = p_{breed} \times litter \times 0.5 \times S_{juv} =$
`r round(params$leslie_F, 4)`, together with the juvenile-to-adult
transition ($S_{juv}$) and adult female survival ($S_{adult}$), gives a
female 2-stage matrix

$$\mathbf{A} = \begin{bmatrix} 0 & F \\ S_{juv} & S_{adult} \end{bmatrix}$$

whose dominant eigenvalue reproduces the published growth rate exactly:
$\lambda =$ `r round(params$lambda_leslie_validated, 4)` (published: 1.047,
+4.7%/year). This validated structure represents about two years of
juvenile survival before an individual breeds (once folded into $F$, once
in the explicit transition), not one, as the recruitment formula might
suggest at first glance.

```{r fig-leslie-maturation}
knitr::include_graphics(params$fig_leslie_maturation)
```

Extending this validated structure across the same adult-survival range
used throughout this note (`r params$s_range[1]`-`r params$s_range[2]`),
holding juvenile survival at Safi's value, $\lambda$ ranges
`r round(params$leslie_vs_pbr$lambda_min, 2)`-`r round(params$leslie_vs_pbr$lambda_max_val, 2)`:
it `r if (params$leslie_vs_pbr$reaches_120) "just reaches" else "does not reach"`
the `r params$thresholds$facility[which.min(params$thresholds$threshold)]` benchmark (1.20) only at the top of the adult-survival range, and
`r if (params$leslie_vs_pbr$reaches_124) "reaches" else "never reaches"`
the `r params$thresholds$facility[which.max(params$thresholds$threshold)]` benchmark (1.24) anywhere in it. Delaying maturation further (more
pre-reproductive stages) pushes $\lambda$ down further still. Independently
of the PBR/Niel-Lebreton analysis above, this validated, published,
species-specific demographic model does not support the imposed
thresholds' implicit growth-rate assumption except at the most optimistic
end of the plausible adult-survival range.

# PVA-lite: 25-year stochastic projection

The Leslie matrix above is deterministic: it describes the average
trajectory but not the year-to-year variability real populations
experience, nor the practical question of whether the currently-imposed
thresholds are likely to be associated with population decline over a
realistic planning horizon. A simplified stochastic population viability
analysis ("PVA-lite") addresses both, without requiring specialised PVA
software: the validated Leslie structure above is projected
`r params$pva_n_years` years forward, redrawing survival and reproduction
each year from Beta distributions (`r params$pva_vital_rate_cv * 100`%
coefficient of variation around Safi's point values -- a working
assumption, since no inter-annual variance estimate for *V. murinus* vital
rates was located) and applying demographic stochasticity (binomial and
Poisson draws on the actual, finite population each year), replicated
`r format(params$pva_n_reps, big.mark = ",")` times per scenario. The
starting population is $N_{min}=$ `r format(params$n0_total, big.mark = ",")`
(both sexes, the local population assumed per facility throughout this
note), split by sex and, within females, across stages using the Leslie
matrix's own stable stage distribution. Three scenarios are compared: no
additional mortality (the model's own baseline trend), and each facility's
imposed threshold applied as extra annual removal, split evenly across
sexes and, within females, proportionally across stages.

```{r fig-pva-projection}
knitr::include_graphics(params$fig_pva_projection)
```

```{r pva-risk-table}
knitr::kable(
  params$pva_risk_summary, digits = c(0, 0, 0, 0, 1, 1),
  col.names = c("Scenario", "Median final N", "P5 final N", "P95 final N", "% decline over horizon", "% quasi-extinction"),
  caption = paste0("Risk summary after ", params$pva_n_years, " years, ",
                    format(params$pva_n_reps, big.mark = ","),
                    " replicate trajectories per scenario. Quasi-extinction: falling below ",
                    round(params$quasi_ext_threshold), " individuals (",
                    100 * (params$quasi_ext_threshold / params$n0_total), "% of N0) at any point.")
)
```

Under this model, the currently-imposed thresholds do not produce a
catastrophic outcome: median population size still grows over
`r params$pva_n_years` years under both. But they measurably raise the
probability of decline relative to the model's own unharvested baseline --
from `r round(params$pva_risk_summary$p_decline[params$pva_risk_summary$scenario == "No additional mortality"], 1)`%
to
`r round(min(params$pva_risk_summary$p_decline[params$pva_risk_summary$scenario != "No additional mortality"]), 0)`-`r round(max(params$pva_risk_summary$p_decline[params$pva_risk_summary$scenario != "No additional mortality"]), 0)`%
-- particularly at `r params$thresholds$facility[which.max(params$thresholds$threshold)]`'s higher threshold, and with a small but non-zero
probability of quasi-extinction there.

# Comparing approaches

## Comparing the three approaches

```{r comparison-table}
knitr::kable(
  params$comparison_summary,
  caption = "Summary comparison: PBR/Niel-Lebreton, the validated species-specific Leslie matrix, and the PVA-lite projection."
)
```

The three methods do not agree. The PBR/Niel-Lebreton approach, at its
fixed benchmarks, sits at the upper end (and, once the recovery factor and
demographic ranges are corrected as in the sections above, potentially
above) what a flexible parametrisation of that same approximation
supports. The validated, published, species-specific Leslie matrix
produces a materially lower growth rate ($\lambda \approx$
`r round(params$lambda_leslie_validated, 3)`) than either PBR benchmark,
reaching them only at the most optimistic edge of the adult-survival
range. The PVA-lite projection, built on that same validated structure,
shows the imposed thresholds are not obviously unsustainable over a
25-year horizon, but they do measurably raise the probability of decline
relative to no additional mortality. Taken together, this three-way
comparison does not resolve which figure is "correct" -- each method
carries its own assumptions and uncertainties, documented in the sections
above -- but it does show that the imposed thresholds sit closer to the
optimistic end of what the available evidence supports than to its
centre, across every method tried, reinforcing the case for treating them
as a band under adaptive management rather than as a precise, fixed limit.

## Inverse demographic boundary analysis

The Leslie matrix cross-check above shows that Safi's (2006) empirical
vital rates for *`r params$species_name`* produce
$\lambda \approx$ `r round(params$lambda_leslie_validated, 3)`, below the
PBR benchmarks. This $\lambda$ is calculated from observed, average-year
vital rates ($\lambda_{observed}$), which is conceptually distinct from
$\lambda_{max}$, the PBR framework's theoretical ceiling under favourable,
non-resource-limited conditions. Falling short of the benchmarks at Safi's
own values therefore does not by itself show that $\lambda_{max} =$
1.20-1.24 is biologically unreachable for this species; what it can show is
which combinations of vital rates the validated structure would need in
order to reach those benchmarks, and how favourable those combinations
would be relative to the empirical baseline.

Holding three of the validated structure's four vital rates (adult female
survival, juvenile survival, breeding fraction, litter size) at Safi's
baseline and solving for the value the fourth would need to reach each PBR
benchmark gives:

```{r leslie-breakeven-table}
knitr::kable(
  params$leslie_breakeven_tbl, digits = c(0, 2, 2, 2, 2, 0, 0),
  col.names = c("Vital rate", "lambda target", "Safi baseline", "Required value",
                "Plausible upper bound", "% of plausible range", "Within plausible range?"),
  caption = "Value each vital rate would need to reach on its own, holding the other three at Safi's baseline, to reach each PBR benchmark. 'NA' marks a target unreachable even at the vital rate's own plausible ceiling."
)
```

No single vital rate reaches $\lambda = 1.20$ within its own plausible
range while the other three stay at Safi's baseline: adult female survival
would need to approach its own upper bound almost exactly; juvenile
survival and litter size would both need to exceed their plausible upper
bounds; and the breeding fraction cannot reach $\lambda = 1.20$ at all,
even at 100% of adult females breeding, reflecting its comparatively low
leverage on $\lambda$ in this structure.

Allowing all four vital rates to vary jointly and independently across
their plausible ranges (a `r format(params$leslie_boundary_range_s_adult[1], nsmall=2)`-`r format(params$leslie_boundary_range_s_adult[2], nsmall=2)`
adult survival range, `r params$leslie_boundary_range_s_juv[1]`-`r params$leslie_boundary_range_s_juv[2]`
for juvenile survival, `r params$leslie_boundary_range_p_breed[1]`-`r params$leslie_boundary_range_p_breed[2]`
for breeding fraction and `r params$leslie_boundary_range_litter[1]`-`r params$leslie_boundary_range_litter[2]`
for litter size; n_stages fixed at 2, the validated structure), only
`r round(params$leslie_boundary_band_shares$pct[params$leslie_boundary_band_shares$band == "1.20-1.24"] + params$leslie_boundary_band_shares$pct[params$leslie_boundary_band_shares$band == ">= 1.24"], 1)`%
of that joint plausible space reaches $\lambda \geq 1.20$, of which
`r round(params$leslie_boundary_band_shares$pct[params$leslie_boundary_band_shares$band == ">= 1.24"], 1)`%
reaches $\lambda \geq 1.24$. Safi's own empirical baseline exceeds
`r round(params$leslie_boundary_pct_below_baseline, 1)`% of this same joint
plausible-space grid.

```{r fig-leslie-boundary-surface}
knitr::include_graphics(params$fig_leslie_boundary_surface)
```

Reaching the PBR benchmarks under the validated demographic structure
therefore requires either a single vital rate pushed to, or beyond, the top
of its plausible range, or several vital rates favourable at the same
time. This characterises how favourable the demographic profile would need
to be relative to what has been measured for this species; it does not, on
its own, establish that $\lambda_{max} = 1.20$-$1.24$ is unreachable as a
theoretical ceiling, since $\lambda_{max}$ and $\lambda_{observed}$ answer
different questions.

## Population size (Nmin): a second, largely separate source of uncertainty

The demographic ceiling ($\lambda_{max}$) explored above and the local
population size ($N_{min}$) it is multiplied by are two separate
components of the PBR equation, carrying separate uncertainties. The
boundary analysis above holds regardless of $N_{min}$, since it concerns a
growth rate, not an abundance; $N_{min}$, in contrast, enters
$PBR = \frac{1}{2}(\lambda_{max}-1) \times F_r \times N_{min}$ linearly, so
any uncertainty in $N_{min}$ carries through to PBR at the same proportion.

No independent population estimate exists for `r params$species_common_name`
in this region, only the assumption reverse-engineered earlier in this note
($N_{min} \approx$ `r format(params$nmin_assumed, big.mark = ",")`,
Reverse-engineering the imposed thresholds). Frick et al. (2026) define
$N_{min}$ explicitly at the scale of an individual facility's own local
population, not at the scale of the species' full range or of a
migratory/regional population: "While standard PBR should be applied at
the scale of an entire population, there are no regulatory mechanisms to
report or respond at that scale. Therefore, fatality thresholds need to be
set on a per-project basis." Their own $N_{min}$ calculation follows the
same logic: "We estimate $N_U$ and $N_L$ for local populations using proxy
bat density estimates applied at a spatial scale relevant to a wind energy
facility," via $N_{local} = \text{Project area (km}^2\text{)} \times 2
\times \text{density estimate}$ -- the same formula this note's own
plausibility check (Reverse-engineering the imposed thresholds) already
applies at each facility's own footprint.

Frick et al. (2026) explicitly flag this project-level scope as a
limitation of their own framework: "we recognize that our approach does
not fully address cumulative impacts at range-wide scales and that future
efforts to devise monitoring and regulatory mechanisms beyond
project-level mitigation are warranted." This caveat is directly relevant
to a regional cluster of nearby facilities such as `r params$project_ref`
(`r params$interproject_distance_km` km apart): if the species' regional
movements link the areas around both facilities, each facility's own
$N_{min}$ and PBR budget, calculated independently as above, would not
capture fatalities occurring at both facilities drawing on the same
underlying population. This note does not attempt to define or quantify an
alternative, regional-population $N_{min}$ in place of the project-level
estimate used throughout -- Frick et al.'s own method provides no basis for
doing so, and no independent population or movement data exist for this
species in this region to constrain such an estimate. The point is
recorded here as a documented, author-acknowledged limitation of the PBR
framework as published, to be weighed alongside the demographic
uncertainty above when the imposed thresholds are discussed with lenders
and other stakeholders.

## Relative importance of demographic capacity and population size

The analyses above distinguish two largely independent sources of
uncertainty in the PBR calculation. The first concerns demographic
capacity: the value of $\lambda_{max}$ that characterises the maximum
potential rate of population growth under favourable conditions,
explored above through the Niel-Lebreton sensitivity analysis, the
species-specific Leslie matrix, the PVA-lite projection and the inverse
demographic boundary analysis. The second concerns scale: the size of the
population, $N_{min}$, to which that demographic capacity is applied,
discussed in the previous section. $\lambda_{max}$ determines a
*proportional* growth capacity; $N_{min}$ converts that proportion into an
*absolute* number of removable individuals. Neither term alone determines
the fatality threshold -- both are required, and $PBR = \frac{1}{2}
(\lambda_{max}-1) \times F_r \times N_{min}$ is exactly linear in each.

Considerable attention is commonly given to $\lambda_{max}$, since it
represents the biological basis for the allowable-removal rate, and the
sections above show that assumption deserves scrutiny for this species.
This does not, however, mean $\lambda_{max}$ is the calculation's larger
source of uncertainty in practice. Frick et al. (2026) state directly:
"The $N_{min}$ value has the most influence on PBR for Least Concern
species" -- directly relevant here, since `r params$species_common_name`
is assessed as Least Concern throughout this note ($F_r=$ `r params$fr_corrected`).

A direct comparison illustrates the same point numerically:

```{r nmin-vs-lambda-sensitivity-table}
knitr::kable(
  params$nmin_vs_lambda_sensitivity, digits = c(0, 0, 0, 0),
  col.names = c("Change", "PBR before", "PBR after", "% change"),
  caption = "PBR's sensitivity to a plausible doubling of Nmin versus the full width of the paper's own fixed lambda_max benchmark range, all else held constant."
)
```

A plausible doubling of $N_{min}$ changes PBR by
`r round(params$nmin_vs_lambda_sensitivity$pct_change[1])`%; moving across
the paper's entire fixed $\lambda_{max}$ benchmark range (1.20 to 1.24)
changes it by only
`r round(params$nmin_vs_lambda_sensitivity$pct_change[2])`%. This is not
because $\lambda_{max}$ is unimportant -- it is because the paper's own
benchmark range is narrow relative to how unconstrained $N_{min}$ is in
this analysis, not because $N_{min}$ is inherently more uncertain in
principle.

```{r fig-nmin-lambda-surface}
knitr::include_graphics(params$fig_nmin_lambda_surface)
```

This has a practical implication for how a PBR threshold should be
periodically reviewed. $\lambda_{max}$ can reasonably be treated as a
relatively stable, species-level demographic benchmark: there is little
reason to revise it often, and it should change mainly in response to
improved demographic evidence (survival, reproduction, age at first
breeding) rather than to fluctuations in annual fatality counts.
$N_{min}$, in contrast, is a locally defined population proxy under this
framework (Population size (Nmin), above) and is the term most likely to
change as better information becomes available on abundance, density,
seasonal movement, migration, and connectivity between facilities --
consistent with Frick et al.'s own framing of adaptive management, which
anticipates updating thresholds as local population density, seasonal bat
activity, or IUCN status information improves. Adaptive refinement of PBR
for `r params$species_common_name` may therefore depend more on improving
the definition and estimation of the population exposed to project
mortality than on further narrowing $\lambda_{max}$ within the range
considered in this note.

# Integrated interpretation and implications

The analyses presented in this note do not identify a single alternative
fatality threshold that can be considered demonstrably more biologically
correct than the currently imposed values. Rather, they show that the
apparent precision of fixed annual thresholds is greater than warranted
by the available demographic and population information.

Within the PBR/Niel-Lebreton framework, and after correcting the recovery
factor to the species' Least Concern status and allowing demographic
parameters to vary across literature-supported ranges, the currently
imposed thresholds fall towards the conservative end of the resulting PBR
distribution. This remains relevant: the existing values should not be
interpreted as uniquely determined biological limits, and alternative
plausible PBR parametrisations produce materially different thresholds.

The additional demographic analyses, however, qualify any simple
inference that the thresholds should therefore be increased. The
validated species-specific Leslie model produces an empirical growth rate
of approximately `r round(params$lambda_leslie_validated, 3)`,
substantially below the fixed PBR $\lambda_{max}$ benchmarks of
1.20-1.24, while the inverse boundary analysis shows that reaching those
benchmarks requires comparatively favourable combinations of vital rates.
The PVA-lite similarly indicates that the current thresholds are not
associated with catastrophic population outcomes under the model
assumptions, but that sustained annual removal at those levels measurably
increases the probability of population decline relative to the
no-additional-mortality scenario.

A further and potentially larger uncertainty concerns $N_{min}$. The
value of approximately `r format(params$nmin_assumed, big.mark = ",")`
explored in this note is a reverse-engineered scenario consistent with
particular PBR assumptions, not an independently estimated population
size. Because $N_{min}$ enters PBR linearly, uncertainty over the size,
spatial extent and connectivity of the population exposed to project
mortality can have greater influence on the numerical threshold than
refinement of $\lambda_{max}$ within the fixed range considered here.
This is particularly relevant for a migratory species and for nearby
facilities that may interact with the same regional population.

Taken together, these results support retaining PBR as a precautionary
**management reference**, but not interpreting the currently imposed
values as precise biological boundaries separating acceptable from
unacceptable mortality. Approaching or exceeding a nominal threshold
should therefore trigger structured review rather than predetermine a
single mitigation response. Such review should consider:

- the magnitude and persistence of exceedance;
- the position of observed mortality relative to the PBR decision range and the assumptions generating that range;
- temporal trends in fatality estimates;
- uncertainty associated with PCFM-derived mortality estimates;
- the conservation relevance and demographic sensitivity of the affected species;
- the spatial concentration and recurrence of fatalities;
- the expected effectiveness and proportionality of mitigation measures;
- evidence of population connectivity, given the PBR framework's own acknowledged limitation for cumulative, range-wide impacts;
- and the broader adaptive-management context in which the threshold is applied.

The principal opportunity for future refinement is therefore not simply
to select a different fixed $\lambda_{max}$, but to progressively improve
the biological interpretation of the population exposed to mortality and
to use accumulating monitoring evidence within an adaptive-management
framework. Where those data remain uncertain, PBR is more appropriately
represented as a decision range with explicitly documented assumptions
than as a uniquely determined annual mortality limit.

# Towards operational adaptive mortality management

The preceding sections address a different question from the one that
actually has to be answered operationally. They ask what fatality
threshold the PBR framework, and the demographic evidence available for
this species, can defensibly support. Operating a facility under PCFM,
however, requires deciding *when*, *where*, and *by how much* to adjust
curtailment, given that fatalities accumulate continuously through the
season and mitigation can be changed while the season is still under way.
An annual threshold checked once at year-end cannot inform that decision
-- by the time an exceedance is confirmed, the mortality has already
occurred and the season is over.

This section proposes three deliberately separated layers, each updated
on a different cadence:

- a **biological reference** (the PBR-derived annual figure discussed
  throughout this note) -- relatively stable, revisited only when
  materially new demographic or population evidence becomes available,
  not adjusted in response to normal year-to-year fatality variation;
- a **dynamic mortality assessment** -- updated weekly from PCFM, tracking
  observed mortality against what the biological reference implies for
  that point in the season;
- a **dynamic, turbine-level curtailment prescription** -- the
  operational output of the assessment above, adjusted as frequently as
  the monitoring data support.

Separating these layers matters for how this is presented to lenders and
regulators: the proposal is not a "flexible PBR" that changes whenever
mortality becomes inconvenient. The biological reference stays fixed; it
is the monitoring-triggered operational response that is dynamic, on a
pre-agreed, documented, and auditable basis.

## From an annual reference to a weekly control chart

Mortality risk for this species is not spread evenly across the year --
it concentrates in the migration windows. Comparing cumulative PCFM
mortality against a flat, linear fraction of the annual threshold
therefore misrepresents how much of the year's risk has already passed.
Instead, the annual reference is converted into an **expected cumulative
trajectory** that follows the same seasonal shape mortality itself
follows, with statistical control limits around it (a Poisson-based
control chart, standard practice for monitoring count data): if realised
mortality is on the reference's own pace, the cumulative count at week
$t$ is approximately Poisson-distributed around the expected value at
that point. Observed cumulative mortality crossing that trajectory
(AMBER) or its upper control limit (RED) is an early-warning signal, not
merely "a bad week", with a full season's worth of response time still
available.

Each trigger tier maps to an escalation step on the turbines already
identified as critical: maintain the current prescription (GREEN);
increase curtailment intensity, e.g. a lower cut-in speed or extended
curtailed hours (AMBER); escalate to a wider turbine subset and the most
conservative cut-in speed for the remainder of the high-risk window, and
trigger a structured, documented review (RED). This is an escalation
ladder, not yet a continuously optimised rule, because only one
before/after curtailment data point currently exists (critical turbines
curtailed vs. not). A fitted dose-response relationship (fatality rate
against cut-in speed, or against % of capacity curtailed) becomes
possible once curtailment intensity itself, not just its presence or
absence, has been varied across enough weeks -- a natural extension once
more seasons of PCFM under varying prescriptions accumulate.

## A first real operational example (2025-2026 PCFM data)

The mechanism above is applied here to the actual carcass search data
available for `r params$project_ref` -- *V. murinus* records from each
project's carcass search log, corrected for detection probability using
the project's working multiplicative rule
(unbiased estimate = `r params$real_correction_factor` x raw carcasses
found). Curtailment at the turbines already identified as critical began
`r params$real_curtailment_start`. Two limitations should be read
alongside everything below: the seasonal risk profile ($q_t$) is derived
from a single pre-curtailment year (2025) at each facility, since that is
the only full year available, so it reflects that year's particular
timing as well as the species' underlying phenology; and the
`r params$real_correction_factor`x rule is a project-level working
assumption, not a fitted, search-effort-specific GenEst model, so the
corrected counts below should be read as indicative, not exact.

```{r fig-real-dashboard}
knitr::include_graphics(params$fig_real_dashboard)
```

```{r real-overage-table}
knitr::kable(
  params$real_overage_summary, digits = c(0, 0, 0, 0, 0),
  col.names = c("Project", "Annual reference", "Raw carcasses (2026 to date)", "GenEst-corrected estimate", "% of annual reference"),
  caption = paste0("2026 year-to-date V. murinus mortality (through ISO week ", params$real_checkpoint_week,
                    ") against the annual PBR reference, both projects.")
)
```

Both projects' 2026 corrected estimate already exceeds the annual
reference for the year to date, driven overwhelmingly by the weeks before
curtailment began (ISO week `r params$real_curtailment_week`): the figure
above shows both projects crossing into the RED zone during the
pre-curtailment spring peak, well before the control chart's own upper
limit would otherwise have been reached later in the season. This is the
concrete version of the point made earlier in this note: an annual figure
checked only at year-end cannot inform an in-season response, because by
the time the exceedance is visible in an annual total, the season -- and
the opportunity to act on it -- is largely over. It is also a genuine,
not hypothetical, illustration of why the reference should not be
interpreted as a hard pass/fail line: exceeding it does not by itself
mean mitigation failed; the pattern of *when* mortality occurred relative
to when curtailment started is what is actually informative here.

```{r real-yoy-table}
knitr::kable(
  params$real_yoy_summary, digits = c(0, 0, 0, 0),
  col.names = c("Project", "2025 (same weeks)", "2026 (same weeks)", "% change"),
  caption = paste0("Year-over-year comparison, GenEst-corrected V. murinus mortality through the same ISO week (",
                    params$real_checkpoint_week, ") in each year.")
)
```

```{r real-yoy-inline}
yoy1 <- params$real_yoy_summary$pct_change[params$real_yoy_summary$project == params$real_overage_summary$project[1]]
yoy2 <- params$real_yoy_summary$pct_change[params$real_yoy_summary$project == params$real_overage_summary$project[2]]
yoy_asymmetry_note <- if (yoy2 >= 0) {
  "an increase, the opposite direction from the first project, and the more actionable signal of the two: it suggests the current prescription at that project has not yet delivered a comparable improvement"
} else if (abs(yoy2) < abs(yoy1) / 2) {
  "also a reduction, but markedly smaller in magnitude than the first project's -- worth watching, since it suggests the current prescription is delivering a smaller benefit at that project"
} else {
  "a reduction of broadly comparable magnitude to the first project's"
}
```

Read alongside the year-to-date total, the year-over-year comparison
gives a more differentiated picture. `r params$real_overage_summary$project[1]`'s
corrected mortality through the same point in the season is
`r round(abs(yoy1))`% `r if (yoy1 < 0) "lower" else "higher"`
than the equivalent period in 2025, consistent with curtailment having a
material effect even though the annual reference was still exceeded.
`r params$real_overage_summary$project[2]`'s corrected mortality through
the same point is `r round(abs(yoy2))`%
`r if (yoy2 < 0) "lower" else "higher"` than in 2025 --
`r yoy_asymmetry_note`, which is exactly the kind of asymmetry between
facilities that a shared annual figure, looked at only in aggregate,
would not surface.

The forward-looking forecast (projected year-end total under the current,
post-curtailment regime continuing unchanged) is included for
completeness, but should be treated cautiously here: it is estimated from
only `r max(params$real_forecast_summary$post_curt_weeks_observed)` weeks
of post-curtailment data so far, too short a window for the estimate of
current-regime intensity to be precise -- the year-over-year comparison
above, which uses no such extrapolation, is the more reliable signal at
this stage. This forecast should firm up, and become genuinely decision-useful,
after a few more weeks of post-curtailment PCFM accumulate.

```{r fig-real-turbine-pareto}
knitr::include_graphics(params$fig_real_turbine_pareto)
```

Turbine identifiers are anonymised (rank-based labels) consistent with
the rest of this note. The distribution of pre-curtailment mortality
across turbines is less concentrated than a simplified illustrative
example might suggest, at either project:
`r params$real_n_turbines_by_project$n_turbines_top80[1]` of
`r params$real_n_turbines_by_project$n_turbines_focus[1]` turbines at
`r params$real_n_turbines_by_project$project[1]`
(`r round(100 * params$real_n_turbines_by_project$n_turbines_top80[1] / params$real_n_turbines_by_project$n_turbines_focus[1])`%
of the fleet), and
`r params$real_n_turbines_by_project$n_turbines_top80[2]` of
`r params$real_n_turbines_by_project$n_turbines_focus[2]` turbines at
`r params$real_n_turbines_by_project$project[2]`
(`r round(100 * params$real_n_turbines_by_project$n_turbines_top80[2] / params$real_n_turbines_by_project$n_turbines_focus[2])`%),
are needed to account for 80% of pre-curtailment mortality, rather than
a small handful at either site. This matters operationally: a strategy
of curtailing only a small minority of turbines is less likely to
capture most of the risk at either project than turbine-level PCFM data
from a more spatially concentrated site would suggest, and
broader-coverage curtailment, informed by this ranking, is likely to be
necessary rather than optional.

The turbine counts here are not directly comparable to the "68 assessed"
/ "27 assessed" figures in the next section: this figure pools every
turbine with at least one recorded *V. murinus* carcass across a longer
window (all of 2025 plus 2026 pre-curtailment), whereas the official
selection universe below is fixed to the single baseline season the
lender-agreed methodology used (April-August 2025 at
`r params$real_n_turbines_by_project$project[1]`, March-August 2025 at
`r params$real_n_turbines_by_project$project[2]`). The longer window here
is a strict superset of the official one -- every officially-assessed
turbine appears in it, plus a handful more that only recorded a carcass
outside that narrower season.

## Turbine-level retrospective validation

The turbines actually curtailed since 5 May 2026 were not selected by this
note's own methodology: they were selected under a separate, lender-agreed
procedure, documented in the facilities' Blanket Curtailment Plan technical
note (May 2026), that ranked turbines by GenEst-estimated *V. murinus*
mortality over a defined baseline season -- April-August 2025 at
`r params$real_overage_summary$project[1]`, March-August 2025 at
`r params$real_overage_summary$project[2]` -- and selected turbines in
ranked order up to 50% of that season's cumulative estimated mortality.
This produced 17 selected turbines at
`r params$real_overage_summary$project[1]` (of 68 assessed) and 7 at
`r params$real_overage_summary$project[2]` (of 27 assessed).

A natural question is how well a simpler, independent approach -- ranking
turbines by raw carcass counts and this note's own
`r params$real_correction_factor`x working rule, rather than a full
search-effort-specific GenEst fit -- would have reproduced that selection,
applied to the same baseline season and the same 50%-cumulative rule. Two
data-coverage caveats apply to this replication and are not smoothed over:
the raw carcass records available run from 22 April 2025 at
`r params$real_overage_summary$project[1]` and from 9 April 2025 at
`r params$real_overage_summary$project[2]`, so the replication is missing
roughly the first three weeks of the nominal April window at the former,
and essentially the entire nominal March window at the latter.

```{r turbine-validation-overlap-table}
knitr::kable(
  params$turbine_validation_overlap,
  digits = c(0, 0, 0, 0, 0, 2, 0, 0),
  col.names = c("Project", "Officially selected", "Independently selected", "Selected by both",
                "Union", "Jaccard index", "% of official selection recovered",
                "% of official selection's mortality captured"),
  caption = "Turbine-level retrospective validation: independent replication (raw carcasses, working correction rule) against the official, lender-agreed selection."
)
```

```{r fig-turbine-validation}
knitr::include_graphics(params$fig_turbine_validation)
```

The independent replication recovers most, but not all, of the officially
selected turbines at both projects, and captures a larger share of their
combined estimated mortality than of their count alone (turbines it
misses tend to be lower-mortality ones near the 50% cutoff, where small
count differences flip a turbine across the line). The remaining gap is
consistent with two compounding factors, not a single cause: the shorter
data window noted above, and the fact that the real, project-specific
GenEst correction is not a flat multiplier -- the implied ratio between
the official estimated mortality and the matching raw carcass count
ranges from `r round(params$turbine_validation_correction_ratio_range[1], 1)`x
to `r round(params$turbine_validation_correction_ratio_range[2], 1)`x
across individual turbines (median
`r round(params$turbine_validation_correction_ratio_median, 1)`x, close to
but visibly more variable than this note's flat
`r params$real_correction_factor`x working assumption), reflecting
genuine differences in search effort, scavenging rates or carcass
persistence between turbines that a project-level average cannot capture.
Practically, this means a simple carcass-count ranking is a reasonable
first approximation for identifying priority turbines between formal
selection exercises, but should not substitute for a full, search-effort-
specific GenEst analysis when a selection decision carries lender or
financial consequences.

## The May-August 2026 emergency response in a full operational year

The analysis above (year-to-date mortality, the year-over-year comparison,
the forward forecast) evaluates the current season against the annual
reference and against 2025 at the same calendar point. A related but
distinct question is how effective the curtailment response has been on
its own terms, and how that effectiveness would play out across a full
"operational year" running from the curtailment start date through the
same date the following year -- explicitly framed, as this note's
adaptive-management proposal is throughout, as evaluating what began as
an emergency response so it can be refined into a longer-term strategy.

The real PCFM records available for this analysis currently run to
`r params$curtailment_year_data_max_date` --
`r params$curtailment_year_checkpoint_week` weeks into the operational
year. What follows should still be read as a checkpoint rather than a
final verdict on the emergency response, and is the first thing to
refresh as later 2026 records continue to accumulate, particularly once
the season's autumn/winter tail is covered.

```{r fig-curtailment-year}
knitr::include_graphics(params$fig_curtailment_year)
```

```{r curtailment-year-reduction-table}
knitr::kable(
  params$curtailment_year_reduction_so_far %>%
    dplyr::select(project, threshold, weeks_of_real_data, actual_corrected_total,
                   counterfactual_corrected_total, avoided_fatalities, pct_reduction),
  digits = c(0, 0, 0, 0, 0, 0, 0),
  col.names = c("Project", "Annual reference", "Weeks of real 2026 data", "Actual (with curtailment)",
                "Counterfactual (no curtailment, 2025 pattern)", "Avoided fatalities", "% reduction"),
  caption = "Mortality avoided so far, over the operational weeks with real 2026 data available, against a no-curtailment counterfactual built from 2025's pattern shifted one year forward."
)
```

`r params$curtailment_year_reduction_so_far$project[1]`
`r if (params$curtailment_year_reduction_so_far$pct_reduction[1] >= 0) "shows a reduction of" else "shows an increase of"`
`r round(abs(params$curtailment_year_reduction_so_far$pct_reduction[1]))`%
(`r round(abs(params$curtailment_year_reduction_so_far$avoided_fatalities[1]))`
fatalities `r if (params$curtailment_year_reduction_so_far$pct_reduction[1] >= 0) "avoided against" else "in excess of"`
the no-curtailment counterfactual), consistent with the year-over-year
comparison above.
`r params$curtailment_year_reduction_so_far$project[2]`
`r if (params$curtailment_year_reduction_so_far$pct_reduction[2] >= 0) "shows a reduction of" else "shows an increase of"`
`r round(abs(params$curtailment_year_reduction_so_far$pct_reduction[2]))`%
(`r round(abs(params$curtailment_year_reduction_so_far$avoided_fatalities[2]))`
fatalities `r if (params$curtailment_year_reduction_so_far$pct_reduction[2] >= 0) "avoided against" else "in excess of"`
the counterfactual) over the same window. With
`r params$curtailment_year_reduction_so_far$weeks_of_real_data[1]` weeks of
real 2026 data now behind this comparison -- most of a growing season,
not just its opening weeks -- this is a materially firmer read than an
early-season snapshot would give, though it should still be read
alongside the year-over-year comparison above rather than in isolation,
and will keep firming up as the operational year's remaining weeks
accumulate.

```{r curtailment-year-counterfactual-table}
knitr::kable(
  params$curtailment_year_full_counterfactual,
  digits = c(0, 0, 0, 0),
  col.names = c("Project", "Annual reference", "Full-year no-curtailment counterfactual", "% of annual reference"),
  caption = "What the full operational year's mortality would total under 2025's pattern, had curtailment never been introduced."
)
```

Extended across the full operational year, the no-curtailment
counterfactual would reach
`r round(params$curtailment_year_full_counterfactual$pct_of_threshold[1] / 100, 1)`x
the annual reference at
`r params$curtailment_year_full_counterfactual$project[1]` and
`r round(params$curtailment_year_full_counterfactual$pct_of_threshold[2] / 100, 1)`x
at `r params$curtailment_year_full_counterfactual$project[2]` -- consistent
with, and independently constructed from, the pre-curtailment overage
already documented in the Blanket Curtailment Plan technical note itself.
This framing also makes explicit a residual-risk point noted earlier in
this note: the current curtailment window does not span the full
operational year, so calendar months outside it remain at, or close to,
counterfactual-level risk under the plan as currently scoped -- the
concrete reason the next iteration of this strategy should extend
monitoring and, where justified by the data, curtailment coverage beyond
the current window, rather than treat the present emergency measure as a
finished design.

## 2027 mitigation frontier: candidate management references

The sections above establish two things separately: Section 5.4 shows
that PBR is far more sensitive to Nmin than to $\lambda_{max}$, and the
Monte Carlo simulation behind Sections 2.6 and 4 propagates uncertainty
in $s$ and $\alpha$ only, at a single, project-shared Nmin held fixed at
`r format(params$nmin_assumed, big.mark = ",")` (the reverse-engineered
working value from "Reverse-engineering the imposed thresholds"). This
section closes that gap and connects the result to the operational
turbine-level and mortality-budgeting evidence already built in this
part of the note. Two things this section is explicitly NOT: a proposal
to change the imposed annual thresholds, and a claim that a wider
plausible PBR range makes 112/88 "wrong". It is a decision-support
comparison, matching the same caution this note has applied throughout
to the relationship between PBR's own uncertainty and any operational
response to it.

### Adding Nmin uncertainty to the Monte Carlo

Nmin is redrawn here from each project's own footprint
(`r params$global_nmin_area_km2` km2) times a plausible density range
(`r params$global_density_low`-`r params$global_density_high`
bats/km2, the same *Pipistrellus pipistrellus* proxy already used for
the Nmin plausibility check in Section 2.4), rather than held fixed --
projects are treated in isolation, not pooled into the
`r format(params$combined_region_km2, big.mark = ",")` km2 combined-region
scenario discussed elsewhere in this note, which remains a separate,
unresolved question.

```{r fig-global-pbr-uncertainty}
knitr::include_graphics(params$fig_global_pbr_uncertainty)
```

```{r global-pbr-old-new-table}
knitr::kable(
  params$global_pbr_old_new_comparison, digits = c(0, 0, 0, 0),
  col.names = c("Scenario", "Q05", "Median", "Q95"),
  caption = "PBR distribution quantiles (bats/year), before and after adding project-specific Nmin uncertainty."
)
```

Adding Nmin uncertainty widens the distribution substantially and shifts
its centre well above the imposed thresholds: the median moves from
`r round(params$global_pbr_old_new_comparison$median[1])` (demographic
uncertainty only) to
`r round(params$global_pbr_old_new_comparison$median[2])`-`r round(params$global_pbr_old_new_comparison$median[3])`
once Nmin is allowed to vary. Under this wider distribution, the
`r params$global_pbr_quantiles_by_project$threshold[1]` threshold at
`r params$global_pbr_quantiles_by_project$project[1]` sits at the
`r round(params$global_pbr_quantiles_by_project$threshold_percentile[1], 1)`th
percentile, and the
`r params$global_pbr_quantiles_by_project$threshold[2]` threshold at
`r params$global_pbr_quantiles_by_project$project[2]` sits at the
`r round(params$global_pbr_quantiles_by_project$threshold_percentile[2], 1)`th
percentile -- both firmly in the conservative tail. As established
throughout this note, this is a read of where the current thresholds sit
within a distribution whose dominant uncertainty (Nmin) is not an
independent population estimate; it is not evidence the thresholds
should move.

### What each candidate level would imply

For each project, five candidate management-reference levels -- the
current imposed threshold, and the Q05, Q25, median and Q75 of that
project's own global PBR distribution above -- are compared on
identical terms: where the level sits in the distribution; the Nmin it
implies at that project's median simulated $\lambda_{max}$; the PVA-lite
decline and quasi-extinction risk of applying that level as a constant
annual removal (same validated Leslie structure and starting population,
$N_0 =$ `r format(params$nmin_assumed, big.mark = ",")`, as Section 4 --
so directly comparable to the risk figures already reported there, not
a new population-size assumption); and the mortality-budgeting
consequence -- the reduction required from the no-curtailment
counterfactual annual total (Section 7.4), and how many turbines,
ranked by pre-curtailment historical contribution (Section 7.2), a
`r round(100 * params$candidate_pbr_effectiveness)`% per-turbine
curtailment-effectiveness assumption (Adams, Gulka & Williams 2021, the
same benchmark the Blanket Curtailment Plan technical note itself cites
to justify its 6.5 m/s cut-in speed) would need to cover.

```{r fig-candidate-pbr}
knitr::include_graphics(params$fig_candidate_pbr)
```

```{r candidate-pbr-table}
knitr::kable(
  params$candidate_pbr_table %>%
    dplyr::mutate(turbines = paste0(n_turbines_required, " of ", n_turbines_available)) %>%
    dplyr::select(project, candidate, level, percentile_in_distribution, implied_nmin_at_median_lambda,
                  pva_p_decline, required_abs_reduction, pct_reduction_needed, turbines),
  digits = c(0, 0, 0, 1, 0, 1, 0, 0, 0),
  col.names = c("Project", "Candidate level", "Bats/yr", "Percentile", "Implied Nmin",
                "PVA: P(decline), %", "Reduction needed", "% reduction", "Turbines needed"),
  caption = "Candidate management references compared on identical terms, per project."
)
```

```{r candidate-pbr-inline}
proj1 <- params$real_overage_summary$project[1]
proj2 <- params$real_overage_summary$project[2]
cand_median_1 <- params$candidate_pbr_table %>% dplyr::filter(candidate == "Median", project == proj1)
cand_median_2 <- params$candidate_pbr_table %>% dplyr::filter(candidate == "Median", project == proj2)
cand_current_1 <- params$candidate_pbr_table %>% dplyr::filter(candidate == "Current imposed threshold", project == proj1)
cand_current_2 <- params$candidate_pbr_table %>% dplyr::filter(candidate == "Current imposed threshold", project == proj2)
```

The starkest result is the PVA risk column, not the turbine counts. At
the median candidate level, 25-year decline probability reaches
`r round(cand_median_1$pva_p_decline)`% at `r proj1` and
`r round(cand_median_2$pva_p_decline)`% at `r proj2` -- far higher than
at the current threshold. This is not a contradiction of the wider
distribution above; it is the direct consequence of the median level
being calibrated to a *larger* implied Nmin
(`r format(round(cand_median_1$implied_nmin_at_median_lambda), big.mark = ",")`)
than the PVA's own starting population, held fixed throughout this note
at the more conservative
`r format(params$nmin_assumed, big.mark = ",")`. Put plainly: moving the
management reference toward the centre of the plausible PBR range is
only as safe as the assumption that the population is actually that
much larger -- an assumption this note has not established, and
precisely the reason the Nmin axis, not $\lambda_{max}$, is the
question that would need to be resolved before any such move could be
considered.

The turbine counts in this table are a like-for-like comparison ACROSS
candidate levels under one uniform assumption, not a substitute for the
turbines actually selected and curtailed. At the current threshold, they
are noticeably larger than the real, already-executed selection
(`r cand_current_1$n_turbines_required` of `r cand_current_1$n_turbines_available`
modelled vs. 17 of 68 actually selected at `r proj1`;
`r cand_current_2$n_turbines_required` of `r cand_current_2$n_turbines_available`
modelled vs. 7 of 27 actually selected at `r proj2`, Sections 7.2-7.3) --
a concrete, quantified confirmation that the real selection was never
dimensioned to bring full-year mortality under the annual reference; it
targeted a defined 50% reduction in a baseline season as an emergency
measure. Reconciling that gap -- what it would actually take, in
turbines and mitigation coverage, to sit reliably under the current
reference across a full year -- is the mortality-budgeting exercise for
the 2027 season this table is built to inform, not a recommendation
to curtail 48 turbines outright: the real next step is testing whether a
smaller, better-targeted set (informed by a second baseline season, and
by which turbines the current 17/7 have already shown to be effective)
can close most of the gap at a materially lower operational cost than
this uniform ranking implies.

## What this shows, and what still needs to happen

This first pass demonstrates that the adaptive-management mechanism
proposed above runs end to end on real data, and already produces
genuinely informative, facility-differentiated signals -- not merely a
proof of concept. It also surfaces concrete next steps rather than
resolving everything: a second pre-curtailment year (or more) would
substantially firm up the seasonal risk profile now based on a single
year at each facility; a longer post-curtailment record would sharpen the
forward-looking forecast and make it decision-useful in its own right,
not just a secondary check on the year-over-year comparison; and two
extensions remain deliberately unimplemented until enough data exist to
justify them rather than to demonstrate them in the abstract --
incorporating energy-loss cost into any future turbine-level allocation
exercise (so a prescription would minimise energy foregone subject to
meeting a required reduction, rather than only ranking by expected
mortality), and an escalate/relax hysteresis rule for adjusting
curtailment down again (requiring several consecutive low-mortality
assessments, with adequate monitoring power to make a low count
informative, before relaxing -- avoiding a rule that escalates and
relaxes on ordinary week-to-week variation).

The two checks above add a further, complementary layer: the turbine
validation shows that an independent, project-level replication of the
real turbine selection is directionally sound but not a substitute for
the real, search-effort-specific analysis behind a lender-facing
decision; and the operational-year comparison, now resting on
`r params$curtailment_year_reduction_so_far$weeks_of_real_data[1]` weeks
of real 2026 data, gives a season-scale read on whether the emergency
response is reducing mortality relative to the no-curtailment
counterfactual at each project -- while also making explicit that the
current curtailment window leaves part of the operational year
uncovered. Both checks should keep being re-run as later 2026 records
accumulate, particularly once the season's autumn/winter tail is
covered, before either is treated as a final verdict on the strategy's
effectiveness.
