---
title: "PBR Interpretation based on Frick et al. 2026 (`r params$project_ref`)"
output:
  word_document:
    toc: true
params:
  project_ref: ""
  species_name: ""
  species_common_name: ""
  thresholds: !r NULL
  fr_original: !r NULL
  fr_corrected: !r NULL
  iucn_status_original_guess: ""
  iucn_status_corrected: ""
  backsolve_grid: !r NULL
  nmin_assumed: !r NULL
  project_footprint_km2: !r NULL
  combined_region_km2: !r NULL
  interproject_distance_km: !r NULL
  nmin_plausibility: !r NULL
  density_proxy_low: !r NULL
  density_proxy_high: !r NULL
  pbr_corrected_fr: !r NULL
  s_range: !r NULL
  alpha_range: !r NULL
  alpha_range_prior: !r NULL
  elasticity_ranges: !r NULL
  fig_elasticity_survival: ""
  fig_elasticity_alpha: ""
  fig_response_surfaces: ""
  fig_response_surfaces_3d: !r NULL
  fig_pbr_density: ""
  n_sim: !r NULL
  pbr_quantiles: !r NULL
  threshold_percentiles: !r NULL
  alpha_range_comparison: !r NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
# knitr::include_graphics() defaults to rewriting absolute paths as
# relative to rmarkdown::render()'s output_dir (here "outputs/", not this
# .Rmd's own "report/" folder) -- pandoc then can't resolve the result and
# silently swaps in the image's alt-text description instead of embedding
# it. The figure paths passed in via params are already absolute; keep
# them as-is.
options(knitr.graphics.rel_path = FALSE)
```

Document history / List of changes.

| Version | Changes | Date of changes | Authors | Review | Review Date | Approval |
|---|---|---|---|---|---|---|
| Version 1 | First draft | 2026-05-07 | Paulo Cardoso | | | |
| Version 2 | Corrected recovery factor (Fr); reverse-engineered and plausibility-checked the local population estimate (Nmin); updated age-at-first-breeding range from a literature review; rebuilt the sensitivity, response-surface and Monte Carlo analyses on consistent, reproducible parameter ranges | 2026-09 | Paulo Cardoso | | | |

# Context

This technical note reviews the Potential Biological Removal (PBR)
framework proposed by Frick et al. (2026) for bats at wind-energy
facilities, and applies it to the `r params$species_common_name` (*`r params$species_name`*)
fatality thresholds currently imposed at `r params$project_ref`
(`r paste(sprintf("%s bats/year at %s", params$thresholds$threshold, params$thresholds$facility), collapse = "; ")`).
No documented derivation of these threshold values is available, and no
independent local population estimate for the species in this region
exists or can realistically be obtained. The note therefore works
backwards from the imposed thresholds to identify a parsimonious,
checkable set of assumptions consistent with them, and then tests how
sensitive those thresholds are to reasonable demographic uncertainty,
rather than to the single fixed parameter values the framework uses by
default.

# PBR Threshold Uncertainty

## Frick et al. adjusted PBR framework

The PBR framework presented by Frick and colleagues (Frick et al. 2026,
herein referred to as Frick) adapts the classical Potential Biological
Removal concept to bat mortality at wind-energy facilities. It retains the
precautionary structure of the original PBR formulation while addressing
known limitations of applying marine-mammal-derived PBR methods to volant
terrestrial taxa, particularly long-lived bat species characterised by low
fecundity, delayed maturity, and strong sensitivity to adult survival.

Under the adjusted framework, PBR is expressed as:

$$PBR = \frac{1}{2} R_{max} \cdot F_r \cdot N_{min}$$

where $R_{max} = \lambda_{max} - 1$ is the maximum annual population growth
rate, $F_r$ is a recovery factor linked to IUCN conservation status, and
$N_{min}$ is a conservative lower-bound estimate of local abundance.

A key conceptual shift in the Frick proposal is the move from broad
regional or continental population framing towards a project-scale local
population proxy. Local abundance bounds ($N_L$ and $N_U$) are derived from
the interaction area surrounding the wind facility combined with plausible
density estimates, and a lognormal formulation is then used to derive the
geometric mean abundance estimate ($\hat{N}$), its coefficient of variation
($CV_{\hat{N}}$), and ultimately $N_{min}$. Uncertainty is therefore
incorporated explicitly into the abundance term: poorly constrained
abundance estimates are penalised with lower, more precautionary PBR
values.

$\lambda_{max}$ is standardised using the Niel & Lebreton (2005)
demographic-invariant approximation, from adult survival ($s$) and age at
first breeding ($\alpha$):

$$\lambda_{max} \approx \frac{(s\alpha - s + \alpha + 1) + \sqrt{(s - s\alpha - \alpha - 1)^2 - 4s\alpha^2}}{2\alpha}$$

Frick et al. propose precautionary fixed $\lambda_{max}$ values (1.24 for
most bats, 1.20 for families with later age at first breeding such as
Pteropodidae) to reduce sensitivity to poorly constrained demographic
inputs. The framework remains structurally dependent on several uncertain
ecological assumptions, including the definition of the local interacting
population, density estimation, recovery-factor selection, and the
demographic assumptions underlying the fixed $\lambda_{max}$ values.
Accordingly, PBR thresholds are best interpreted as precautionary decision
bands rather than deterministic operational limits, particularly where
they are used to calibrate mitigation measures such as curtailment.

## Sensitivity and elasticity analysis

Sensitivity and elasticity analyses were undertaken to evaluate how
uncertainty in the two demographic parameters used by the Niel & Lebreton
approximation, adult survival ($s$) and age at first breeding ($\alpha$),
propagates into $\lambda_{max}$ and consequently into PBR. Parameter ranges
were set at $s \in$ [`r params$s_range[1]`, `r params$s_range[2]`] and
$\alpha \in$ [`r params$alpha_range[1]`, `r params$alpha_range[2]`] years,
the latter following the literature review below. For each combination,
$\lambda_{max}$ was recalculated and local sensitivity estimated
numerically using finite differences. Elasticity is a dimensionless
proportional sensitivity metric: the expected percentage variation in
$\lambda_{max}$ resulting from a 1% variation in a demographic parameter.

```{r elasticity-table}
knitr::kable(params$elasticity_ranges, digits = 3,
             caption = "Elasticity of lambda_max to each demographic parameter.")
```

```{r elasticity-inline}
es <- params$elasticity_ranges[params$elasticity_ranges$Parameter == "Adult survival (s)", ]
ea <- params$elasticity_ranges[params$elasticity_ranges$Parameter == "Age at first breeding (alpha)", ]
```

$\lambda_{max}$ is substantially more sensitive to uncertainty in adult
survival than to uncertainty in age at first breeding across the explored
demographic space: a 1% variation in adult survival produces roughly
`r round(abs(es$min), 2)`-`r round(abs(es$max), 2)`% variation in
$\lambda_{max}$, versus `r round(abs(ea$min), 2)`-`r round(abs(ea$max), 2)`%
for an equivalent variation in age at first breeding. Sensitivity is not
spatially uniform: combinations of relatively high adult survival and
earlier reproductive maturity produce the largest elastic responses, which
is particularly relevant because adult survival is also among the least
well-constrained demographic parameters for most bat species.

```{r fig-elasticity-survival}
knitr::include_graphics(params$fig_elasticity_survival)
```

```{r fig-elasticity-alpha}
knitr::include_graphics(params$fig_elasticity_alpha)
```

These results support the central rationale of the Frick adjusted
framework: uncertainty in demographic parameterisation can materially
influence resulting PBR values, and precautionary fixed-$\lambda_{max}$
assumptions reduce excessive sensitivity to uncertain demographic inputs.
However, substantial uncertainty remains even under a fixed-$\lambda_{max}$
approach, through the interaction of demographic assumptions, local
abundance estimation, and recovery-factor selection -- the remainder of
this note works through that interaction for the `r params$species_common_name`
at `r params$project_ref` specifically.

## Evidence base for age at first breeding

For PBR, $\alpha$ should represent effective age at first reproduction --
the age at which a female actually recruits offspring into the population
-- not merely physiological sexual maturity. A literature review for
*`r params$species_name`* was undertaken on this basis.

| Source | Species / group | Relevant information | Implication for alpha |
|---|---|---|---|
| AnAge database | *Vespertilio murinus* | Female and male sexual maturity listed at 730 days (~2 years); litter size 2; one litter per year. | Supports alpha ~= 2 as a defensible value. |
| Haensel 2010, *Nyctalus* | *Vespertilio murinus* | Some females raise young in their first year of life; the author's own data suggest most or all females may reach sexual maturity in the year of birth and raise young as one-year-olds, stated cautiously. | Suggests alpha = 1-2 may be plausible; alpha = 2 is not an optimistic assumption. |
| Ratpenats species account | *Vespertilio murinus* | Many females mate in their first autumn. | Supports early sexual maturity, but mating is not the same as confirmed breeding. |
| Racey & Entwistle synthesis, cited in Frick et al.'s own discussion | Bats broadly | The general view for most bats is that age at first breeding is around 1 year, though first-year breeding probability can be lower; Frick et al. use alpha = 1.5 as a general benchmark and suggest alpha = 2 for some families. | Supports alpha = 1.5-2 as plausible for many bat species. |
| Cryan et al. 2012 | Migratory tree bats killed at wind turbines | Ovarian development indicating sexual maturity even in first-year females for several migratory tree-roosting species. | Useful analogue: some wind-vulnerable migratory bats mature very early. |
| Komar et al. 2020 | Male *Vespertilio murinus* | Food restriction delayed seasonal sexual maturation in males; well-fed males matured earlier. | Maturation timing is condition-dependent; male data are less directly useful for PBR (which uses female recruitment). |

Table: Sources reviewed for age at first breeding in the Parti-coloured
bat.

$\alpha = 2$ is a defensible, reasonably conservative central value for PBR
purposes under the effective-recruitment reading of $\alpha$, and is
directly supported by the AnAge sexual-maturity estimate. $\alpha = 1.5$
(the paper's own general bat benchmark) is a plausible lower bound, though
it rests on thinner, more cautiously stated species-specific evidence.
$\alpha = 3-4$ has no direct empirical support for *V. murinus* in the
sources reviewed; it remains a useful precautionary sensitivity scenario
for ecological constraints on effective recruitment (migration costs, poor
juvenile survival, low first-year breeding probability), but is not the
best-supported biological estimate. This review supports using
$\alpha \in$ [`r params$alpha_range[1]`, `r params$alpha_range[2]`] years as
the working range throughout this note, in place of a single fixed value.

## Reverse-engineering the imposed thresholds

`r params$project_ref` `r if (nrow(params$thresholds) > 1) "were" else "was"` given
a fixed annual fatality threshold for the `r params$species_common_name` --
`r paste(sprintf("%s bats/year at %s", params$thresholds$threshold, params$thresholds$facility), collapse = " and ")`
-- without a documented derivation. Since no independent local population
estimate for the species exists for this region, plausibility is assessed
by working backwards from the imposed thresholds under the Frick PBR
equation.

```{r backsolve-table}
knitr::kable(params$backsolve_grid, digits = c(0, 0, 2, 0),
             col.names = c("Facility", "Threshold (bats/yr)", "Fixed lambda_max", "Required Nmin"),
             caption = paste0("Local population size (Nmin) required to reproduce each imposed threshold, under Fr = ",
                               params$fr_original, " (", params$iucn_status_original_guess,
                               ") and the paper's fixed lambda_max benchmarks."))
```

This is the most parsimonious explanation available for how the imposed
thresholds were obtained: local populations of
~`r format(params$nmin_assumed, big.mark = ",")` individuals near each
facility, combined with the paper's $\lambda_{max}$ benchmarks applied per
facility. `r params$project_ref` are approximately
`r params$interproject_distance_km` km apart, so this is better read as two
similarly sized local populations rather than a single population shared
between them (see below).

Whether ~`r format(params$nmin_assumed, big.mark = ",")` individuals is a
plausible local population can be assessed against the real, known project
geography, by asking what average bat density it would imply, rather than
backing out a hypothetical area from a hypothetical density.

```{r nmin-plausibility-table}
knitr::kable(
  params$nmin_plausibility, digits = c(0, 0, 2, 2),
  col.names = c("Area scenario", "Area (km2)", "Implied density low (bats/km2)", "Implied density high (bats/km2)"),
  caption = paste0("Density that Nmin = ", format(params$nmin_assumed, big.mark = ","),
                    " would imply, under the two real area readings, holding the density ratio fixed at the default proxy's (",
                    params$density_proxy_low, "-", params$density_proxy_high, " bats/km2).")
)
```

The single-project-footprint reading
(`r params$nmin_plausibility$area_km2[1]` km$^2$) implies a density of
`r round(params$nmin_plausibility$density_low[1], 1)`-`r round(params$nmin_plausibility$density_high[1], 1)`
bats/km$^2$ -- closely matching the paper's own default density proxy
(`r params$density_proxy_low`-`r params$density_proxy_high` bats/km$^2$,
based on *Pipistrellus pipistrellus*, used in the absence of
species-specific density data) for a common bat species. Treating
$N_{min} \approx `r format(params$nmin_assumed, big.mark = ",")`$ as the
local population near each facility individually, rather than shared
across both, is therefore consistent with typical bat density and with
Frick et al.'s own project-scale definition of local population.

By contrast, treating
$N_{min} = `r format(params$nmin_assumed, big.mark = ",")`$ as a single
population spread across the combined region spanning both projects
(`r params$nmin_plausibility$area_km2[2]` km$^2$) implies a density of only
`r round(params$nmin_plausibility$density_low[2], 2)`-`r round(params$nmin_plausibility$density_high[2], 2)`
bats/km$^2$ -- roughly an order of magnitude below the default proxy, an
assumption that would need independent supporting evidence (e.g. that
regional density for `r params$species_common_name` here is genuinely that
sparse) before being considered reasonable.

The project-footprint reading is therefore both the more defensible order
of magnitude and the one that matches Frick et al.'s own project-scale
local-population concept. The working assumption throughout the rest of
this note is accordingly that $N_{min} \approx `r format(params$nmin_assumed, big.mark = ",")`$
applies to each facility's own local population, not to a single population
shared across the `r params$interproject_distance_km` km separating
`r paste(params$thresholds$facility, collapse = " and ")`.

$F_r = `r params$fr_original`$ corresponds to a `r params$iucn_status_original_guess`
status. The `r params$species_common_name` is not a threatened category, so
the defensible recovery factor is $F_r = `r params$fr_corrected`$
(`r params$iucn_status_corrected`). Holding $N_{min}$ at
`r format(params$nmin_assumed, big.mark = ",")` and correcting only $F_r$:

```{r pbr-corrected-fr-table}
knitr::kable(params$pbr_corrected_fr, digits = c(2, 0),
             col.names = c("Fixed lambda_max", "PBR (bats/yr)"),
             caption = paste0("PBR at Nmin = ", format(params$nmin_assumed, big.mark = ","),
                               ", Fr = ", params$fr_corrected, "."))
```

The recovery-factor correction alone shifts PBR upward at every fixed
$\lambda_{max}$ benchmark, before any change to $\lambda_{max}$ itself is
considered.

## PBR response surfaces

```{r fig-response-surfaces}
knitr::include_graphics(params$fig_response_surfaces)
```

```{r fig-response-surfaces-3d}
if (!is.null(params$fig_response_surfaces_3d)) {
  knitr::include_graphics(params$fig_response_surfaces_3d)
}
```

Across the plausible demographic range, the imposed-threshold contours
only appear for a narrow combination of relatively high adult survival and
low age at first breeding -- outside that narrow band, PBR departs from
the imposed thresholds even before changing $F_r$.

## Monte Carlo simulation outputs

Uncertainty in adult survival and age at first breeding was propagated
into $\lambda_{max}$ and PBR by simulation, holding
$N_{min} = `r format(params$nmin_assumed, big.mark = ",")`$ and
$F_r = `r params$fr_corrected`$ fixed, with
$s \sim U(`r params$s_range[1]`, `r params$s_range[2]`)$ and
$\alpha \sim U(`r params$alpha_range[1]`, `r params$alpha_range[2]`)$
(`r format(params$n_sim, big.mark = ",")` draws).

```{r mc-quantile-table}
q_tbl <- tibble::tibble(
  Quantile = names(params$pbr_quantiles),
  `PBR (bats/yr)` = round(params$pbr_quantiles, 1)
)
knitr::kable(q_tbl, caption = "Simulated PBR distribution quantiles.")
```

```{r mc-percentile-table}
knitr::kable(
  params$threshold_percentiles, digits = 2,
  col.names = c("Facility", "Threshold (bats/yr)", "Percentile rank in simulated PBR", "Median ratio", "P95 ratio"),
  caption = "Where the imposed thresholds fall within the simulated PBR distribution."
)
```

The imposed thresholds sit on the low, conservative side of the simulated
distribution, not the permissive side: median simulated PBR is
`r round(min(params$threshold_percentiles$median_ratio), 2)`-`r round(max(params$threshold_percentiles$median_ratio), 2)`x
the imposed thresholds, reaching
`r round(min(params$threshold_percentiles$p95_ratio), 2)`-`r round(max(params$threshold_percentiles$p95_ratio), 2)`x
at the 95th percentile.

```{r mc-alpha-comparison-table}
alpha_comp_names <- c(
  "Alpha range", "Q05", "Median", "Q95",
  sprintf("%% below %s (%s)", params$thresholds$threshold, params$thresholds$facility)
)
knitr::kable(
  params$alpha_range_comparison, digits = 1, col.names = alpha_comp_names,
  caption = paste0("Effect of the alpha range on the simulated PBR distribution (Fr = ",
                    params$fr_corrected, ", Nmin = ", format(params$nmin_assumed, big.mark = ","), ").")
)
```

Moving from the wider, less literature-grounded alpha range
(`r params$alpha_range_prior[1]`-`r params$alpha_range_prior[2]`) to the
range the evidence review supports
(`r params$alpha_range[1]`-`r params$alpha_range[2]`) shifts the simulated
PBR distribution further up, not down: the literature review therefore
reinforces, rather than weakens, the conclusion below.

```{r fig-pbr-density}
knitr::include_graphics(params$fig_pbr_density)
```

## Preliminary interpretation and implications

Under a defensible, non-optimistic parametrisation -- recovery factor
corrected to the species' actual (non-threatened) status, and demographic
parameters treated as a literature-supported range rather than a single
fixed point -- the currently imposed thresholds sit near the conservative
end of the resulting PBR distribution rather than at its centre. This
indicates the current thresholds are not unreasonably permissive; if
anything, a flexible parametrisation would support materially higher
values, approaching double the current figures at the upper end of the
plausible range.

These results reinforce that any operational threshold derived from PBR
should be accompanied by transparent documentation of the underlying
calculations, demographic assumptions, local population estimates, and
uncertainty analysis. A nominal threshold can still serve as a
precautionary reference point, but mitigation decisions should
additionally consider:

- the plausible uncertainty interval surrounding the threshold;
- temporal trends in fatality estimates;
- uncertainty associated with PCFM-derived mortality estimates;
- the conservation relevance and demographic sensitivity of the affected species;
- the spatial concentration and recurrence of fatalities;
- the expected effectiveness and proportionality of mitigation measures;
- and the broader adaptive-management context in which the threshold is applied.

Presenting PBR as a range rather than a single fixed number provides a
more robust and defensible basis for adaptive management decisions where
demographic parameters, local population structure, and abundance
estimates remain incompletely resolved -- as is the case for the
`r params$species_common_name` in this region.
