---
title: "PBR Interpretation based on Frick et al. 2026 (`r params$project_ref`)"
output:
  word_document:
    toc: true
params:
  project_ref: ""
  species_name: ""
  species_common_name: ""
  thresholds: !r NULL
  fr_original: !r NULL
  fr_corrected: !r NULL
  iucn_status_original_guess: ""
  iucn_status_corrected: ""
  backsolve_grid: !r NULL
  nmin_assumed: !r NULL
  project_footprint_km2: !r NULL
  combined_region_km2: !r NULL
  interproject_distance_km: !r NULL
  nmin_plausibility: !r NULL
  density_proxy_low: !r NULL
  density_proxy_high: !r NULL
  pbr_corrected_fr: !r NULL
  s_range: !r NULL
  alpha_range: !r NULL
  alpha_range_prior: !r NULL
  elasticity_ranges: !r NULL
  fig_lambda_max: ""
  fig_elasticity_survival: ""
  fig_elasticity_alpha: ""
  fig_response_surfaces: ""
  fig_response_surfaces_3d: !r NULL
  fig_response_surfaces_3d_html: !r NULL
  fig_pbr_density: ""
  n_sim: !r NULL
  pbr_quantiles: !r NULL
  threshold_percentiles: !r NULL
  alpha_range_comparison: !r NULL
  leslie_s_juv: !r NULL
  leslie_s_adult_f: !r NULL
  leslie_s_adult_m: !r NULL
  leslie_p_breed: !r NULL
  leslie_litter: !r NULL
  leslie_F: !r NULL
  lambda_leslie_validated: !r NULL
  leslie_vs_pbr: !r NULL
  fig_leslie_maturation: ""
  n0_total: !r NULL
  pva_n_years: !r NULL
  pva_n_reps: !r NULL
  pva_vital_rate_cv: !r NULL
  quasi_ext_threshold: !r NULL
  pva_risk_summary: !r NULL
  fig_pva_projection: ""
  comparison_summary: !r NULL
  leslie_breakeven_tbl: !r NULL
  leslie_boundary_band_shares: !r NULL
  leslie_boundary_pct_below_baseline: !r NULL
  leslie_boundary_range_s_adult: !r NULL
  leslie_boundary_range_s_juv: !r NULL
  leslie_boundary_range_p_breed: !r NULL
  leslie_boundary_range_litter: !r NULL
  fig_leslie_boundary_surface: ""
  nmin_vs_lambda_sensitivity: !r NULL
  fig_nmin_lambda_surface: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
# knitr::include_graphics() defaults to rewriting absolute paths as
# relative to rmarkdown::render()'s output_dir (here "outputs/", not this
# .Rmd's own "report/" folder) -- pandoc then can't resolve the result and
# silently swaps in the image's alt-text description instead of embedding
# it. The figure paths passed in via params are already absolute; keep
# them as-is.
options(knitr.graphics.rel_path = FALSE)
```

Document history / List of changes.

| Version | Changes | Date of changes | Authors |
|---|---|---|---|
| Version 1 | First draft | 2026-05-07 | Paulo Cardoso |
| Version 2 | Corrected recovery factor (Fr); reverse-engineered and plausibility-checked the local population estimate (Nmin); updated age-at-first-breeding range from a literature review; rebuilt the sensitivity, response-surface and Monte Carlo analyses on consistent, reproducible parameter ranges | 2026-09 | Paulo Cardoso |
| Version 3 | Added a Leslie-matrix cross-check (validated against the published Dekker & Limpens 2024 model) and a 25-year stochastic PVA-lite projection, with a summary comparison against the PBR/Niel-Lebreton approach | 2026-09 | Paulo Cardoso |
| Version 4 | Added an inverse demographic boundary analysis on the validated Leslie matrix, and a discussion of Nmin as a source of uncertainty, including Frick et al.'s own definition of local population size and its stated limitation for cumulative impacts across nearby facilities | 2026-09 | Paulo Cardoso |
| Version 5 | Replaced the kaleido-based 3D response-surface export with a fixed-camera static image; anonymised facility names to Project 1/Project 2 for external presentation; added a direct quantitative comparison of PBR's sensitivity to Nmin versus lambda_max, and a synthesising Nmin x lambda_max surface figure | 2026-09 | Paulo Cardoso |

# Context

This technical note reviews the Potential Biological Removal (PBR)
framework proposed by Frick et al. (2026) for bats at wind-energy
facilities, and applies it to the `r params$species_common_name` (*`r params$species_name`*)
fatality thresholds currently imposed at `r params$project_ref`
(`r paste(sprintf("%s bats/year at %s", params$thresholds$threshold, params$thresholds$facility), collapse = "; ")`).
No documented derivation of these threshold values is available, and no
independent local population estimate for the species in this region
exists or can realistically be obtained. The note therefore works
backwards from the imposed thresholds to identify a parsimonious,
checkable set of assumptions consistent with them, and then tests how
sensitive those thresholds are to reasonable demographic uncertainty,
rather than to the single fixed parameter values the framework uses by
default.

# PBR Threshold Uncertainty

## Frick et al. adjusted PBR framework

The PBR framework presented by Frick and colleagues (Frick et al. 2026,
herein referred to as Frick) adapts the classical Potential Biological
Removal concept to bat mortality at wind-energy facilities. It retains the
precautionary structure of the original PBR formulation while addressing
known limitations of applying marine-mammal-derived PBR methods to volant
terrestrial taxa, particularly long-lived bat species characterised by low
fecundity, delayed maturity, and strong sensitivity to adult survival.

Under the adjusted framework, PBR is expressed as:

$$PBR = \frac{1}{2} R_{max} \cdot F_r \cdot N_{min}$$

where $R_{max} = \lambda_{max} - 1$ is the maximum annual population growth
rate, $F_r$ is a recovery factor linked to IUCN conservation status, and
$N_{min}$ is a conservative lower-bound estimate of local abundance.

A key conceptual shift in the Frick proposal is the move from broad
regional or continental population framing towards a project-scale local
population proxy. Local abundance bounds ($N_L$ and $N_U$) are derived from
the interaction area surrounding the wind facility combined with plausible
density estimates, and a lognormal formulation is then used to derive the
geometric mean abundance estimate ($\hat{N}$), its coefficient of variation
($CV_{\hat{N}}$), and ultimately $N_{min}$. Uncertainty is therefore
incorporated explicitly into the abundance term: poorly constrained
abundance estimates are penalised with lower, more precautionary PBR
values.

$\lambda_{max}$ is standardised using the Niel & Lebreton (2005)
demographic-invariant approximation, from adult survival ($s$) and age at
first breeding ($\alpha$):

$$\lambda_{max} \approx \frac{(s\alpha - s + \alpha + 1) + \sqrt{(s - s\alpha - \alpha - 1)^2 - 4s\alpha^2}}{2\alpha}$$

Frick et al. propose precautionary fixed $\lambda_{max}$ values (1.24 for
most bats, 1.20 for families with later age at first breeding such as
Pteropodidae) to reduce sensitivity to poorly constrained demographic
inputs. The framework remains structurally dependent on several uncertain
ecological assumptions, including the definition of the local interacting
population, density estimation, recovery-factor selection, and the
demographic assumptions underlying the fixed $\lambda_{max}$ values.
Accordingly, PBR thresholds are best interpreted as precautionary decision
bands rather than deterministic operational limits, particularly where
they are used to calibrate mitigation measures such as curtailment.

## Sensitivity and elasticity analysis

Sensitivity and elasticity analyses were undertaken to evaluate how
uncertainty in the two demographic parameters used by the Niel & Lebreton
approximation, adult survival ($s$) and age at first breeding ($\alpha$),
propagates into $\lambda_{max}$ and consequently into PBR. Parameter ranges
were set at $s \in$ [`r params$s_range[1]`, `r params$s_range[2]`] and
$\alpha \in$ [`r params$alpha_range[1]`, `r params$alpha_range[2]`] years,
the latter following the literature review below. For each combination,
$\lambda_{max}$ was recalculated and local sensitivity estimated
numerically using finite differences. Elasticity is a dimensionless
proportional sensitivity metric: the expected percentage variation in
$\lambda_{max}$ resulting from a 1% variation in a demographic parameter.

```{r fig-lambda-max}
knitr::include_graphics(params$fig_lambda_max)
```

Within the explored range, $\lambda_{max}$ *decreases* as adult survival or
age at first breeding increase -- a counterintuitive relationship
documented by Niel & Lebreton (2005) and noted by Frick et al. themselves,
arising from the life-history trade-off between survival and fecundity
built into the demographic-invariant approximation. This is a decrease in
absolute terms, not a sign change: for any survival probability in (0, 1),
this formula's $\lambda_{max}$ only approaches 1 (a stable population) in
the limit as age at first breeding tends to infinity, and never actually
reaches it for a realistic age at first breeding -- the population is
growing ($\lambda_{max} > 1$) throughout the entire range explored here and
well beyond it. The elasticity values below should be read with that in
mind: a negative elasticity means $\lambda_{max}$ falls (while remaining
well above 1) as the parameter increases, not that $\lambda_{max}$ itself
becomes small or negative.

```{r elasticity-table}
knitr::kable(params$elasticity_ranges, digits = 3,
             caption = "Elasticity of lambda_max to each demographic parameter.")
```

```{r elasticity-inline}
es <- params$elasticity_ranges[params$elasticity_ranges$Parameter == "Adult survival (s)", ]
ea <- params$elasticity_ranges[params$elasticity_ranges$Parameter == "Age at first breeding (alpha)", ]
```

$\lambda_{max}$ is substantially more sensitive to uncertainty in adult
survival than to uncertainty in age at first breeding across the explored
demographic space: a 1% variation in adult survival produces roughly
`r round(abs(es$min), 2)`-`r round(abs(es$max), 2)`% variation in
$\lambda_{max}$, versus `r round(abs(ea$min), 2)`-`r round(abs(ea$max), 2)`%
for an equivalent variation in age at first breeding. Sensitivity is not
spatially uniform: combinations of relatively high adult survival and
earlier reproductive maturity produce the largest elastic responses, which
is particularly relevant because adult survival is also among the least
well-constrained demographic parameters for most bat species.

```{r fig-elasticity-survival}
knitr::include_graphics(params$fig_elasticity_survival)
```

```{r fig-elasticity-alpha}
knitr::include_graphics(params$fig_elasticity_alpha)
```

These results support the central rationale of the Frick adjusted
framework: uncertainty in demographic parameterisation can materially
influence resulting PBR values, and precautionary fixed-$\lambda_{max}$
assumptions reduce excessive sensitivity to uncertain demographic inputs.
However, substantial uncertainty remains even under a fixed-$\lambda_{max}$
approach, through the interaction of demographic assumptions, local
abundance estimation, and recovery-factor selection -- the remainder of
this note works through that interaction for the `r params$species_common_name`
at `r params$project_ref` specifically.

## Evidence base for age at first breeding

For PBR, $\alpha$ should represent effective age at first reproduction --
the age at which a female actually recruits offspring into the population
-- not merely physiological sexual maturity. A literature review for
*`r params$species_name`* was undertaken on this basis.

| Source | Species / group | Relevant information | Implication for alpha |
|---|---|---|---|
| AnAge database | *Vespertilio murinus* | Female and male sexual maturity listed at 730 days (~2 years); litter size 2; one litter per year. | Supports alpha ~= 2 as a defensible value. |
| Haensel 2010, *Nyctalus* | *Vespertilio murinus* | Some females raise young in their first year of life; the author's own data suggest most or all females may reach sexual maturity in the year of birth and raise young as one-year-olds, stated cautiously. | Suggests alpha = 1-2 may be plausible; alpha = 2 is not an optimistic assumption. |
| Ratpenats species account | *Vespertilio murinus* | Many females mate in their first autumn. | Supports early sexual maturity, but mating is not the same as confirmed breeding. |
| Racey & Entwistle synthesis, cited in Frick et al.'s own discussion | Bats broadly | The general view for most bats is that age at first breeding is around 1 year, though first-year breeding probability can be lower; Frick et al. use alpha = 1.5 as a general benchmark and suggest alpha = 2 for some families. | Supports alpha = 1.5-2 as plausible for many bat species. |
| Cryan et al. 2012 | Migratory tree bats killed at wind turbines | Ovarian development indicating sexual maturity even in first-year females for several migratory tree-roosting species. | Useful analogue: some wind-vulnerable migratory bats mature very early. |
| Komar et al. 2020 | Male *Vespertilio murinus* | Food restriction delayed seasonal sexual maturation in males; well-fed males matured earlier. | Maturation timing is condition-dependent; male data are less directly useful for PBR (which uses female recruitment). |

Table: Sources reviewed for age at first breeding in the Parti-coloured
bat.

$\alpha = 2$ is a defensible, reasonably conservative central value for PBR
purposes under the effective-recruitment reading of $\alpha$, and is
directly supported by the AnAge sexual-maturity estimate. $\alpha = 1.5$
(the paper's own general bat benchmark) is a plausible lower bound, though
it rests on thinner, more cautiously stated species-specific evidence.
$\alpha = 3-4$ has no direct empirical support for *V. murinus* in the
sources reviewed; it remains a useful precautionary sensitivity scenario
for ecological constraints on effective recruitment (migration costs, poor
juvenile survival, low first-year breeding probability), but is not the
best-supported biological estimate. This review supports using
$\alpha \in$ [`r params$alpha_range[1]`, `r params$alpha_range[2]`] years as
the working range throughout this note, in place of a single fixed value.

## Reverse-engineering the imposed thresholds

`r params$project_ref` `r if (nrow(params$thresholds) > 1) "were" else "was"` given
a fixed annual fatality threshold for the `r params$species_common_name` --
`r paste(sprintf("%s bats/year at %s", params$thresholds$threshold, params$thresholds$facility), collapse = " and ")`
-- without a documented derivation. Since no independent local population
estimate for the species exists for this region, plausibility is assessed
by working backwards from the imposed thresholds under the Frick PBR
equation.

```{r backsolve-table}
knitr::kable(params$backsolve_grid, digits = c(0, 0, 2, 0),
             col.names = c("Facility", "Threshold (bats/yr)", "Fixed lambda_max", "Required Nmin"),
             caption = paste0("Local population size (Nmin) required to reproduce each imposed threshold, under Fr = ",
                               params$fr_original, " (", params$iucn_status_original_guess,
                               ") and the paper's fixed lambda_max benchmarks."))
```

Under $F_r=$ `r params$fr_original` (`r params$iucn_status_original_guess`)
and the paper's fixed $\lambda_{max}$ benchmarks applied per facility, a
local population of ~`r format(params$nmin_assumed, big.mark = ",")`
individuals near each facility reproduces the imposed thresholds exactly.
This shows that ~`r format(params$nmin_assumed, big.mark = ",")` is *one*
parameter combination consistent with the imposed thresholds under those
assumptions -- it is not evidence that a population of that size was
itself the value used when the thresholds were originally derived, since
the parameters actually used for that derivation are not documented and
were not provided along with the thresholds. `r params$project_ref` are
approximately `r params$interproject_distance_km` km apart, so if a local
population of this size is assumed at all, it is better read as two
similarly sized local populations rather than a single population shared
between them (see below).

Whether ~`r format(params$nmin_assumed, big.mark = ",")` individuals is a
plausible local population can be assessed against the real, known project
geography, by asking what average bat density it would imply, rather than
backing out a hypothetical area from a hypothetical density.

```{r nmin-plausibility-table}
knitr::kable(
  params$nmin_plausibility, digits = c(0, 0, 2, 2),
  col.names = c("Area scenario", "Area (km2)", "Implied density low (bats/km2)", "Implied density high (bats/km2)"),
  caption = paste0("Density that Nmin = ", format(params$nmin_assumed, big.mark = ","),
                    " would imply, under the two real area readings, holding the density ratio fixed at the default proxy's (",
                    params$density_proxy_low, "-", params$density_proxy_high, " bats/km2).")
)
```

The single-project-footprint reading
(`r params$nmin_plausibility$area_km2[1]` km$^2$) implies a density of
`r round(params$nmin_plausibility$density_low[1], 1)`-`r round(params$nmin_plausibility$density_high[1], 1)`
bats/km$^2$ -- closely matching the paper's own default density proxy
(`r params$density_proxy_low`-`r params$density_proxy_high` bats/km$^2$,
based on *Pipistrellus pipistrellus*, used in the absence of
species-specific density data) for a common bat species. Treating
$N_{min} \approx `r format(params$nmin_assumed, big.mark = ",")`$ as the
local population near each facility individually, rather than shared
across both, is therefore consistent with typical bat density and with
Frick et al.'s own project-scale definition of local population.

By contrast, treating
$N_{min} = `r format(params$nmin_assumed, big.mark = ",")`$ as a single
population spread across the combined region spanning both projects
(`r params$nmin_plausibility$area_km2[2]` km$^2$) implies a density of only
`r round(params$nmin_plausibility$density_low[2], 2)`-`r round(params$nmin_plausibility$density_high[2], 2)`
bats/km$^2$ -- roughly an order of magnitude below the default proxy, an
assumption that would need independent supporting evidence (e.g. that
regional density for `r params$species_common_name` here is genuinely that
sparse) before being considered reasonable.

The project-footprint reading is therefore both the more defensible order
of magnitude and the one that matches Frick et al.'s own project-scale
local-population concept. The working assumption throughout the rest of
this note is accordingly that $N_{min} \approx `r format(params$nmin_assumed, big.mark = ",")`$
applies to each facility's own local population, not to a single population
shared across the `r params$interproject_distance_km` km separating
`r paste(params$thresholds$facility, collapse = " and ")`.

$F_r = `r params$fr_original`$ corresponds to a `r params$iucn_status_original_guess`
status. The `r params$species_common_name` is not a threatened category, so
the defensible recovery factor is $F_r = `r params$fr_corrected`$
(`r params$iucn_status_corrected`). Holding $N_{min}$ at
`r format(params$nmin_assumed, big.mark = ",")` and correcting only $F_r$:

```{r pbr-corrected-fr-table}
knitr::kable(params$pbr_corrected_fr, digits = c(2, 0),
             col.names = c("Fixed lambda_max", "PBR (bats/yr)"),
             caption = paste0("PBR at Nmin = ", format(params$nmin_assumed, big.mark = ","),
                               ", Fr = ", params$fr_corrected, "."))
```

The recovery-factor correction alone shifts PBR upward at every fixed
$\lambda_{max}$ benchmark, before any change to $\lambda_{max}$ itself is
considered.

## PBR response surfaces

```{r fig-response-surfaces}
knitr::include_graphics(params$fig_response_surfaces)
```

```{r fig-response-surfaces-3d}
if (!is.null(params$fig_response_surfaces_3d)) {
  knitr::include_graphics(params$fig_response_surfaces_3d)
}
```

`r if (!is.null(params$fig_response_surfaces_3d_html)) paste0("A live, rotatable version of this figure (hover for exact values, toggle Fr scenarios) is provided as a standalone file, *", basename(params$fig_response_surfaces_3d_html), "*, alongside this document -- Word cannot embed an interactive plot directly, but the file opens in any browser without needing a server.")`

Across the plausible demographic range, the imposed-threshold contours
only appear for a narrow combination of relatively high adult survival and
low age at first breeding -- outside that narrow band, PBR departs from
the imposed thresholds even before changing $F_r$.

## Monte Carlo simulation outputs

Uncertainty in adult survival and age at first breeding was propagated
into $\lambda_{max}$ and PBR by simulation, holding
$N_{min} = `r format(params$nmin_assumed, big.mark = ",")`$ and
$F_r = `r params$fr_corrected`$ fixed, with
$s \sim U(`r params$s_range[1]`, `r params$s_range[2]`)$ and
$\alpha \sim U(`r params$alpha_range[1]`, `r params$alpha_range[2]`)$
(`r format(params$n_sim, big.mark = ",")` draws).

```{r mc-quantile-table}
q_tbl <- tibble::tibble(
  Quantile = names(params$pbr_quantiles),
  `PBR (bats/yr)` = round(params$pbr_quantiles, 1)
)
knitr::kable(q_tbl, caption = "Simulated PBR distribution quantiles.")
```

```{r mc-percentile-table}
knitr::kable(
  params$threshold_percentiles, digits = 2,
  col.names = c("Facility", "Threshold (bats/yr)", "Percentile rank in simulated PBR", "Median ratio", "P95 ratio"),
  caption = "Where the imposed thresholds fall within the simulated PBR distribution."
)
```

The imposed thresholds sit on the low, conservative side of the simulated
distribution, not the permissive side: median simulated PBR is
`r round(min(params$threshold_percentiles$median_ratio), 2)`-`r round(max(params$threshold_percentiles$median_ratio), 2)`x
the imposed thresholds, reaching
`r round(min(params$threshold_percentiles$p95_ratio), 2)`-`r round(max(params$threshold_percentiles$p95_ratio), 2)`x
at the 95th percentile.

```{r mc-alpha-comparison-table}
alpha_comp_names <- c(
  "Alpha range", "Q05", "Median", "Q95",
  sprintf("%% below %s (%s)", params$thresholds$threshold, params$thresholds$facility)
)
knitr::kable(
  params$alpha_range_comparison, digits = 1, col.names = alpha_comp_names,
  caption = paste0("Effect of the alpha range on the simulated PBR distribution (Fr = ",
                    params$fr_corrected, ", Nmin = ", format(params$nmin_assumed, big.mark = ","), ").")
)
```

Moving from the wider, less literature-grounded alpha range
(`r params$alpha_range_prior[1]`-`r params$alpha_range_prior[2]`) to the
range the evidence review supports
(`r params$alpha_range[1]`-`r params$alpha_range[2]`) shifts the simulated
PBR distribution further up, not down: the literature review therefore
reinforces, rather than weakens, the conclusion below.

```{r fig-pbr-density}
knitr::include_graphics(params$fig_pbr_density)
```

# Leslie matrix cross-check

The PBR/Niel-Lebreton approach is one specific method among several used to
estimate a population's sustainable growth rate: simpler methods (e.g. a
Leslie/Lefkovitch matrix) and more complex ones (full population viability
analysis, PVA) are also standard practice, and checking whether they agree
is itself informative. Unlike Niel & Lebreton's demographic-invariant
approximation, which is deliberately built to need only adult survival and
age at first breeding, a Leslie matrix requires an explicit juvenile
survival rate and fecundity -- inputs the PBR calculation above does not
use at all.

Species-specific values for the Parti-coloured bat were located: Safi
(2006) reports juvenile survival = `r params$leslie_s_juv`, adult female
survival = `r params$leslie_s_adult_f`, adult male survival =
`r params$leslie_s_adult_m`, and `r params$leslie_p_breed * 100`% of adult
females breeding; Zhigalin & Moskvitina (2017) report a litter size of
`r params$leslie_litter` at suburban colonies (2.7-2.9 at urban colonies).
Safi's survival figures are simple return rates (animals marked in year
*t* recaptured in *t+1*), not estimates from a modern open-population
mark-recapture model, and no confidence intervals were reported --
treated here as central working values, not precisely known rates.

This parametrisation was cross-checked against Dekker & Limpens (2024),
"Population dynamics of Dutch bats relevant to wind energy", which
independently built and published a *V. murinus* stage matrix from the
same Safi (2006) and Zhigalin & Moskvitina data. The female recruitment
term they use, $F = p_{breed} \times litter \times 0.5 \times S_{juv} =$
`r round(params$leslie_F, 4)`, together with the juvenile-to-adult
transition ($S_{juv}$) and adult female survival ($S_{adult}$), gives a
female 2-stage matrix

$$\mathbf{A} = \begin{bmatrix} 0 & F \\ S_{juv} & S_{adult} \end{bmatrix}$$

whose dominant eigenvalue reproduces the published growth rate exactly:
$\lambda =$ `r round(params$lambda_leslie_validated, 4)` (published: 1.047,
+4.7%/year). This validated structure represents about two years of
juvenile survival before an individual breeds (once folded into $F$, once
in the explicit transition), not one, as the recruitment formula might
suggest at first glance.

```{r fig-leslie-maturation}
knitr::include_graphics(params$fig_leslie_maturation)
```

Extending this validated structure across the same adult-survival range
used throughout this note (`r params$s_range[1]`-`r params$s_range[2]`),
holding juvenile survival at Safi's value, $\lambda$ ranges
`r round(params$leslie_vs_pbr$lambda_min, 2)`-`r round(params$leslie_vs_pbr$lambda_max_val, 2)`:
it `r if (params$leslie_vs_pbr$reaches_120) "just reaches" else "does not reach"`
the `r params$thresholds$facility[params$thresholds$threshold == 120]` benchmark (1.20) only at the top of the adult-survival range, and
`r if (params$leslie_vs_pbr$reaches_124) "reaches" else "never reaches"`
the `r params$thresholds$facility[params$thresholds$threshold == 144]` benchmark (1.24) anywhere in it. Delaying maturation further (more
pre-reproductive stages) pushes $\lambda$ down further still. Independently
of the PBR/Niel-Lebreton analysis above, this validated, published,
species-specific demographic model does not support the imposed
thresholds' implicit growth-rate assumption except at the most optimistic
end of the plausible adult-survival range.

# Comparing approaches

## PVA-lite: 25-year stochastic projection

The Leslie matrix above is deterministic: it describes the average
trajectory but not the year-to-year variability real populations
experience, nor the practical question of whether the currently-imposed
thresholds are likely to be associated with population decline over a
realistic planning horizon. A simplified stochastic population viability
analysis ("PVA-lite") addresses both, without requiring specialised PVA
software: the validated Leslie structure above is projected
`r params$pva_n_years` years forward, redrawing survival and reproduction
each year from Beta distributions (`r params$pva_vital_rate_cv * 100`%
coefficient of variation around Safi's point values -- a working
assumption, since no inter-annual variance estimate for *V. murinus* vital
rates was located) and applying demographic stochasticity (binomial and
Poisson draws on the actual, finite population each year), replicated
`r format(params$pva_n_reps, big.mark = ",")` times per scenario. The
starting population is $N_{min}=$ `r format(params$n0_total, big.mark = ",")`
(both sexes, the local population assumed per facility throughout this
note), split by sex and, within females, across stages using the Leslie
matrix's own stable stage distribution. Three scenarios are compared: no
additional mortality (the model's own baseline trend), and each facility's
imposed threshold applied as extra annual removal, split evenly across
sexes and, within females, proportionally across stages.

```{r fig-pva-projection}
knitr::include_graphics(params$fig_pva_projection)
```

```{r pva-risk-table}
knitr::kable(
  params$pva_risk_summary, digits = c(0, 0, 0, 0, 1, 1),
  col.names = c("Scenario", "Median final N", "P5 final N", "P95 final N", "% decline over horizon", "% quasi-extinction"),
  caption = paste0("Risk summary after ", params$pva_n_years, " years, ",
                    format(params$pva_n_reps, big.mark = ","),
                    " replicate trajectories per scenario. Quasi-extinction: falling below ",
                    round(params$quasi_ext_threshold), " individuals (",
                    100 * (params$quasi_ext_threshold / params$n0_total), "% of N0) at any point.")
)
```

Under this model, the currently-imposed thresholds do not produce a
catastrophic outcome: median population size still grows over
`r params$pva_n_years` years under both. But they measurably raise the
probability of decline relative to the model's own unharvested baseline --
from `r round(params$pva_risk_summary$p_decline[params$pva_risk_summary$scenario == "No additional mortality"], 1)`%
to
`r round(min(params$pva_risk_summary$p_decline[params$pva_risk_summary$scenario != "No additional mortality"]), 0)`-`r round(max(params$pva_risk_summary$p_decline[params$pva_risk_summary$scenario != "No additional mortality"]), 0)`%
-- particularly at `r params$thresholds$facility[which.max(params$thresholds$threshold)]`'s higher threshold, and with a small but non-zero
probability of quasi-extinction there.

## Comparing the three approaches

```{r comparison-table}
knitr::kable(
  params$comparison_summary,
  caption = "Summary comparison: PBR/Niel-Lebreton, the validated species-specific Leslie matrix, and the PVA-lite projection."
)
```

The three methods do not agree. The PBR/Niel-Lebreton approach, at its
fixed benchmarks, sits at the upper end (and, once the recovery factor and
demographic ranges are corrected as in the sections above, potentially
above) what a flexible parametrisation of that same approximation
supports. The validated, published, species-specific Leslie matrix
produces a materially lower growth rate ($\lambda \approx$
`r round(params$lambda_leslie_validated, 3)`) than either PBR benchmark,
reaching them only at the most optimistic edge of the adult-survival
range. The PVA-lite projection, built on that same validated structure,
shows the imposed thresholds are not obviously unsustainable over a
25-year horizon, but they do measurably raise the probability of decline
relative to no additional mortality. Taken together, this three-way
comparison does not resolve which figure is "correct" -- each method
carries its own assumptions and uncertainties, documented in the sections
above -- but it does show that the imposed thresholds sit closer to the
optimistic end of what the available evidence supports than to its
centre, across every method tried, reinforcing the case for treating them
as a band under adaptive management rather than as a precise, fixed limit.

## Inverse demographic boundary analysis

The Leslie matrix cross-check above shows that Safi's (2006) empirical
vital rates for *`r params$species_name`* produce
$\lambda \approx$ `r round(params$lambda_leslie_validated, 3)`, below the
PBR benchmarks. This $\lambda$ is calculated from observed, average-year
vital rates ($\lambda_{observed}$), which is conceptually distinct from
$\lambda_{max}$, the PBR framework's theoretical ceiling under favourable,
non-resource-limited conditions. Falling short of the benchmarks at Safi's
own values therefore does not by itself show that $\lambda_{max} =$
1.20-1.24 is biologically unreachable for this species; what it can show is
which combinations of vital rates the validated structure would need in
order to reach those benchmarks, and how favourable those combinations
would be relative to the empirical baseline.

Holding three of the validated structure's four vital rates (adult female
survival, juvenile survival, breeding fraction, litter size) at Safi's
baseline and solving for the value the fourth would need to reach each PBR
benchmark gives:

```{r leslie-breakeven-table}
knitr::kable(
  params$leslie_breakeven_tbl, digits = c(0, 2, 2, 2, 2, 0, 0),
  col.names = c("Vital rate", "lambda target", "Safi baseline", "Required value",
                "Plausible upper bound", "% of plausible range", "Within plausible range?"),
  caption = "Value each vital rate would need to reach on its own, holding the other three at Safi's baseline, to reach each PBR benchmark. 'NA' marks a target unreachable even at the vital rate's own plausible ceiling."
)
```

No single vital rate reaches $\lambda = 1.20$ within its own plausible
range while the other three stay at Safi's baseline: adult female survival
would need to approach its own upper bound almost exactly; juvenile
survival and litter size would both need to exceed their plausible upper
bounds; and the breeding fraction cannot reach $\lambda = 1.20$ at all,
even at 100% of adult females breeding, reflecting its comparatively low
leverage on $\lambda$ in this structure.

Allowing all four vital rates to vary jointly and independently across
their plausible ranges (a `r format(params$leslie_boundary_range_s_adult[1], nsmall=2)`-`r format(params$leslie_boundary_range_s_adult[2], nsmall=2)`
adult survival range, `r params$leslie_boundary_range_s_juv[1]`-`r params$leslie_boundary_range_s_juv[2]`
for juvenile survival, `r params$leslie_boundary_range_p_breed[1]`-`r params$leslie_boundary_range_p_breed[2]`
for breeding fraction and `r params$leslie_boundary_range_litter[1]`-`r params$leslie_boundary_range_litter[2]`
for litter size; n_stages fixed at 2, the validated structure), only
`r round(params$leslie_boundary_band_shares$pct[params$leslie_boundary_band_shares$band == "1.20-1.24"] + params$leslie_boundary_band_shares$pct[params$leslie_boundary_band_shares$band == ">= 1.24"], 1)`%
of that joint plausible space reaches $\lambda \geq 1.20$, of which
`r round(params$leslie_boundary_band_shares$pct[params$leslie_boundary_band_shares$band == ">= 1.24"], 1)`%
reaches $\lambda \geq 1.24$. Safi's own empirical baseline exceeds
`r round(params$leslie_boundary_pct_below_baseline, 1)`% of this same joint
plausible-space grid.

```{r fig-leslie-boundary-surface}
knitr::include_graphics(params$fig_leslie_boundary_surface)
```

Reaching the PBR benchmarks under the validated demographic structure
therefore requires either a single vital rate pushed to, or beyond, the top
of its plausible range, or several vital rates favourable at the same
time. This characterises how favourable the demographic profile would need
to be relative to what has been measured for this species; it does not, on
its own, establish that $\lambda_{max} = 1.20$-$1.24$ is unreachable as a
theoretical ceiling, since $\lambda_{max}$ and $\lambda_{observed}$ answer
different questions.

## Population size (Nmin): a second, largely separate source of uncertainty

The demographic ceiling ($\lambda_{max}$) explored above and the local
population size ($N_{min}$) it is multiplied by are two separate
components of the PBR equation, carrying separate uncertainties. The
boundary analysis above holds regardless of $N_{min}$, since it concerns a
growth rate, not an abundance; $N_{min}$, in contrast, enters
$PBR = \frac{1}{2}(\lambda_{max}-1) \times F_r \times N_{min}$ linearly, so
any uncertainty in $N_{min}$ carries through to PBR at the same proportion.

No independent population estimate exists for `r params$species_common_name`
in this region, only the assumption reverse-engineered earlier in this note
($N_{min} \approx$ `r format(params$nmin_assumed, big.mark = ",")`,
Reverse-engineering the imposed thresholds). Frick et al. (2026) define
$N_{min}$ explicitly at the scale of an individual facility's own local
population, not at the scale of the species' full range or of a
migratory/regional population: "While standard PBR should be applied at
the scale of an entire population, there are no regulatory mechanisms to
report or respond at that scale. Therefore, fatality thresholds need to be
set on a per-project basis." Their own $N_{min}$ calculation follows the
same logic: "We estimate $N_U$ and $N_L$ for local populations using proxy
bat density estimates applied at a spatial scale relevant to a wind energy
facility," via $N_{local} = \text{Project area (km}^2\text{)} \times 2
\times \text{density estimate}$ -- the same formula this note's own
plausibility check (Reverse-engineering the imposed thresholds) already
applies at each facility's own footprint.

Frick et al. (2026) explicitly flag this project-level scope as a
limitation of their own framework: "we recognize that our approach does
not fully address cumulative impacts at range-wide scales and that future
efforts to devise monitoring and regulatory mechanisms beyond
project-level mitigation are warranted." This caveat is directly relevant
to a regional cluster of nearby facilities such as `r params$project_ref`
(`r params$interproject_distance_km` km apart): if the species' regional
movements link the areas around both facilities, each facility's own
$N_{min}$ and PBR budget, calculated independently as above, would not
capture fatalities occurring at both facilities drawing on the same
underlying population. This note does not attempt to define or quantify an
alternative, regional-population $N_{min}$ in place of the project-level
estimate used throughout -- Frick et al.'s own method provides no basis for
doing so, and no independent population or movement data exist for this
species in this region to constrain such an estimate. The point is
recorded here as a documented, author-acknowledged limitation of the PBR
framework as published, to be weighed alongside the demographic
uncertainty above when the imposed thresholds are discussed with lenders
and other stakeholders.

## Relative importance of demographic capacity and population size

The analyses above distinguish two largely independent sources of
uncertainty in the PBR calculation. The first concerns demographic
capacity: the value of $\lambda_{max}$ that characterises the maximum
potential rate of population growth under favourable conditions,
explored above through the Niel-Lebreton sensitivity analysis, the
species-specific Leslie matrix, the PVA-lite projection and the inverse
demographic boundary analysis. The second concerns scale: the size of the
population, $N_{min}$, to which that demographic capacity is applied,
discussed in the previous section. $\lambda_{max}$ determines a
*proportional* growth capacity; $N_{min}$ converts that proportion into an
*absolute* number of removable individuals. Neither term alone determines
the fatality threshold -- both are required, and $PBR = \frac{1}{2}
(\lambda_{max}-1) \times F_r \times N_{min}$ is exactly linear in each.

Considerable attention is commonly given to $\lambda_{max}$, since it
represents the biological basis for the allowable-removal rate, and the
sections above show that assumption deserves scrutiny for this species.
This does not, however, mean $\lambda_{max}$ is the calculation's larger
source of uncertainty in practice. Frick et al. (2026) state directly:
"The $N_{min}$ value has the most influence on PBR for Least Concern
species" -- directly relevant here, since `r params$species_common_name`
is assessed as Least Concern throughout this note ($F_r=$ `r params$fr_corrected`).

A direct comparison illustrates the same point numerically:

```{r nmin-vs-lambda-sensitivity-table}
knitr::kable(
  params$nmin_vs_lambda_sensitivity, digits = c(0, 0, 0, 0),
  col.names = c("Change", "PBR before", "PBR after", "% change"),
  caption = "PBR's sensitivity to a plausible doubling of Nmin versus the full width of the paper's own fixed lambda_max benchmark range, all else held constant."
)
```

A plausible doubling of $N_{min}$ changes PBR by
`r round(params$nmin_vs_lambda_sensitivity$pct_change[1])`%; moving across
the paper's entire fixed $\lambda_{max}$ benchmark range (1.20 to 1.24)
changes it by only
`r round(params$nmin_vs_lambda_sensitivity$pct_change[2])`%. This is not
because $\lambda_{max}$ is unimportant -- it is because the paper's own
benchmark range is narrow relative to how unconstrained $N_{min}$ is in
this analysis, not because $N_{min}$ is inherently more uncertain in
principle.

```{r fig-nmin-lambda-surface}
knitr::include_graphics(params$fig_nmin_lambda_surface)
```

This has a practical implication for how a PBR threshold should be
periodically reviewed. $\lambda_{max}$ can reasonably be treated as a
relatively stable, species-level demographic benchmark: there is little
reason to revise it often, and it should change mainly in response to
improved demographic evidence (survival, reproduction, age at first
breeding) rather than to fluctuations in annual fatality counts.
$N_{min}$, in contrast, is a locally defined population proxy under this
framework (Population size (Nmin), above) and is the term most likely to
change as better information becomes available on abundance, density,
seasonal movement, migration, and connectivity between facilities --
consistent with Frick et al.'s own framing of adaptive management, which
anticipates updating thresholds as local population density, seasonal bat
activity, or IUCN status information improves. Adaptive refinement of PBR
for `r params$species_common_name` may therefore depend more on improving
the definition and estimation of the population exposed to project
mortality than on further narrowing $\lambda_{max}$ within the range
considered in this note.

# Preliminary interpretation and implications

Under a defensible, non-optimistic parametrisation -- recovery factor
corrected to the species' actual (non-threatened) status, and demographic
parameters treated as a literature-supported range rather than a single
fixed point -- the currently imposed thresholds sit near the conservative
end of the resulting PBR distribution rather than at its centre. This
indicates the current thresholds are not unreasonably permissive; if
anything, a flexible parametrisation would support materially higher
values, approaching double the current figures at the upper end of the
plausible range.

These results reinforce that any operational threshold derived from PBR
should be accompanied by transparent documentation of the underlying
calculations, demographic assumptions, local population estimates, and
uncertainty analysis. A nominal threshold can still serve as a
precautionary reference point, but mitigation decisions should
additionally consider:

- the plausible uncertainty interval surrounding the threshold;
- temporal trends in fatality estimates;
- uncertainty associated with PCFM-derived mortality estimates;
- the conservation relevance and demographic sensitivity of the affected species;
- the spatial concentration and recurrence of fatalities;
- the expected effectiveness and proportionality of mitigation measures;
- whether nearby facilities may draw on a shared regional or migratory population, given the PBR framework's own acknowledged limitation for cumulative, range-wide impacts;
- and the broader adaptive-management context in which the threshold is applied.

The distinction between demographic capacity and population size drawn in
the preceding section should guide how this threshold is periodically
reviewed: revisiting $N_{min}$ as better abundance, movement and
connectivity evidence becomes available is likely to matter more than
further narrowing $\lambda_{max}$, for the `r params$species_common_name`
in this region.
