---
title: "Bat Mortality Mitigation -- Illustrative Scenario (`r params$project_ref`)"
output:
  word_document:
    toc: false
params:
  project_ref: ""
  species_common_name: ""
  threshold: !r NULL
  n_turbines_total: !r NULL
  n_turbines_curtailed_target: !r NULL
  correction_factor: !r NULL
  nmin_assumed: !r NULL
  pbr_corrected_fr: !r NULL
  turbine_hist: !r NULL
  effectiveness: !r NULL
  summary_tbl: !r NULL
  fig_turbine_pareto: ""
  fig_effectiveness: ""
  turbine_validation_comparison: !r NULL
  turbine_validation_overlap_n: !r NULL
  turbine_validation_overlap_pct: !r NULL
  fig_turbine_validation: ""
  turbine_classification: !r NULL
  turbine_classification_summary: !r NULL
  turbine_classification_core_residual_n: !r NULL
  turbine_classification_candidates: !r NULL
  turbine_classification_weak_pct: !r NULL
  turbine_classification_boundary_value: !r NULL
  fig_turbine_classification: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
options(knitr.graphics.rel_path = FALSE)
```

*Illustrative scenario prepared for the Paris Expert Workshop. `r params$project_ref`,
`r params$n_turbines_total` WTG, and every mortality figure below are
fictional and fully simulated -- none of the numbers in this document
are drawn from, or derived by rescaling, any real project's data. The
method shown (PBR reference, turbine-level targeting, response
effectiveness, adaptive classification) is the one applied to real
projects; only the numbers are synthetic.*

# 1. Reference threshold

An annual mortality reference of `r params$threshold` fatalities/year is
assumed for `r params$species_common_name` at `r params$project_ref`
(`r params$n_turbines_total` WTG), following the PBR framework of Frick
et al. (2026). At the framework's benchmark growth-rate values, this
implies:

```{r pbr-table}
knitr::kable(
  params$pbr_corrected_fr,
  digits = c(2, 1),
  col.names = c("lambda_max benchmark", "Implied PBR")
)
```

Correction of raw carcass counts to estimated mortality uses an
illustrative detection-correction factor of
`r sprintf("%.1f", params$correction_factor)`x throughout.

# 2. Where mortality concentrates

```{r fig-pareto}
knitr::include_graphics(params$fig_turbine_pareto)
```

Ranking turbines by Baseline-period mortality and selecting the minimum
set covering half of cumulative mortality lands on
`r sum(params$turbine_hist$curtailed)` of `r params$n_turbines_total`
turbines -- close to the
`r params$n_turbines_curtailed_target`-turbine target for this scenario,
emerging from the same rule used operationally, not fixed in advance.

```{r fig-validation}
knitr::include_graphics(params$fig_turbine_validation)
```

Because this scenario is simulated, the turbines' true underlying risk is
known and can be compared against the ranking actually used for
curtailment: `r params$turbine_validation_overlap_n` of
`r sum(params$turbine_hist$curtailed)`
(`r params$turbine_validation_overlap_pct`%) of the turbines selected on
observed mortality are also among the true highest-risk turbines --
illustrating that an observed-data ranking recovers most, not all, of the
underlying risk pattern, which is the expected behaviour of the method
rather than a shortcoming of this particular selection.

# 3. Response effectiveness

```{r fig-effectiveness}
knitr::include_graphics(params$fig_effectiveness)
```

```{r summary-table}
knitr::kable(params$summary_tbl, col.names = c("Metric", "Value"))
```

Measured against a no-response counterfactual built from the Baseline
period's own pattern, the response period ends the year below the
annual reference -- the same qualitative result obtained on the real
projects this method was developed for.

# 4. Outlook: next-period turbine classification

```{r fig-classification}
knitr::include_graphics(params$fig_turbine_classification)
```

```{r classification-table}
knitr::kable(params$turbine_classification_summary)
```

Core turbines (already curtailed) show
`r params$turbine_classification_core_residual_n` with a recorded
fatality since the response began -- a signal to review intensification,
never grounds for relaxation. `r length(params$turbine_classification_candidates)`
turbine(s) outside the curtailed set meet the corroboration bar for
expansion (Baseline mortality comparable to the curtailed turbines' and
a confirmed record since the response began). The 2025-only baseline is
never averaged with post-response data, and low post-response mortality
at an already-curtailed turbine is read as the intervention working, not
as evidence that turbine is low-risk.

# What this illustrates

- A single annual reference translates into turbine-level targeting via
  one consistent rule (minimum set covering half of cumulative
  mortality), not a separate judgement call per turbine.
- Observed-data selection recovers most, but rarely all, of the true
  risk ranking -- an expected property of the method, not a defect.
- Response effectiveness is read against an explicit counterfactual, not
  simply "was the annual reference exceeded."
- Expansion candidates require corroboration from both the untreated
  baseline and confirmed post-response recurrence; residual fatalities
  at curtailed turbines are monitored, never treated as grounds to
  relax mitigation there.
