---
title: "PBR Interpretation based on Frick et al. 2026"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})

source(here::here("R", "pbr_functions.R"))
```

```{r source-thresholds, include=FALSE}
# Re-uses R/pbr_frick_bsh_dgy_thresholds.R as-is, so the report and the
# independently runnable script can never drift apart: same seed, same
# simulation, same numbers. Console output and its own plot are suppressed
# here; only the named objects it creates are used below.
old_wd <- setwd(here::here())
source("R/pbr_frick_bsh_dgy_thresholds.R", local = knitr::knit_global())
setwd(old_wd)
```

Document history / List of changes.

| Version | Changes | Date of changes | Authors | Review | Review Date | Approval |
|---|---|---|---|---|---|---|
| Version 1 | First draft | 2026-05-07 | Paulo Cardoso | | | |
| Version 2 | Corrected recovery factor (Fr); reverse-engineered and plausibility-checked the local population estimate (Nmin); updated age-at-first-breeding range from a literature review; rebuilt the sensitivity, response-surface and Monte Carlo analyses on consistent, reproducible parameter ranges | 2026-09 | Paulo Cardoso | | | |

# Context

This technical note reviews the Potential Biological Removal (PBR)
framework proposed by Frick et al. (2026) for bats at wind-energy
facilities, and applies it to the Parti-coloured bat (*Vespertilio
murinus*) fatality thresholds currently imposed at the BSH and DGY wind
farms (144 and 120 bats/year respectively). No documented derivation of
these two threshold values is available, and no independent local
population estimate for the species in this region exists or can
realistically be obtained. The note therefore works backwards from the
imposed thresholds to identify a parsimonious, checkable set of
assumptions consistent with them, and then tests how sensitive those
thresholds are to reasonable demographic uncertainty, rather than to the
single fixed parameter values the framework uses by default.

# PBR Threshold Uncertainty

## Frick et al. adjusted PBR framework

The PBR framework presented by Frick and colleagues (Frick et al. 2026,
herein referred to as Frick) adapts the classical Potential Biological
Removal concept to bat mortality at wind-energy facilities. It retains the
precautionary structure of the original PBR formulation while addressing
known limitations of applying marine-mammal-derived PBR methods to volant
terrestrial taxa, particularly long-lived bat species characterised by low
fecundity, delayed maturity, and strong sensitivity to adult survival.

Under the adjusted framework, PBR is expressed as:

$$PBR = \frac{1}{2} R_{max} \cdot F_r \cdot N_{min}$$

where $R_{max} = \lambda_{max} - 1$ is the maximum annual population growth
rate, $F_r$ is a recovery factor linked to IUCN conservation status, and
$N_{min}$ is a conservative lower-bound estimate of local abundance.

A key conceptual shift in the Frick proposal is the move from broad
regional or continental population framing towards a project-scale local
population proxy. Local abundance bounds ($N_L$ and $N_U$) are derived from
the interaction area surrounding the wind facility combined with plausible
density estimates, and a lognormal formulation is then used to derive the
geometric mean abundance estimate ($\hat{N}$), its coefficient of variation
($CV_{\hat{N}}$), and ultimately $N_{min}$. Uncertainty is therefore
incorporated explicitly into the abundance term: poorly constrained
abundance estimates are penalised with lower, more precautionary PBR
values.

$\lambda_{max}$ is standardised using the Niel & Lebreton (2005)
demographic-invariant approximation, from adult survival ($s$) and age at
first breeding ($\alpha$):

$$\lambda_{max} \approx \frac{(s\alpha - s + \alpha + 1) + \sqrt{(s - s\alpha - \alpha - 1)^2 - 4s\alpha^2}}{2\alpha}$$

Frick et al. propose precautionary fixed $\lambda_{max}$ values (1.24 for
most bats, 1.20 for families with later age at first breeding such as
Pteropodidae) to reduce sensitivity to poorly constrained demographic
inputs. The framework remains structurally dependent on several uncertain
ecological assumptions, including the definition of the local interacting
population, density estimation, recovery-factor selection, and the
demographic assumptions underlying the fixed $\lambda_{max}$ values.
Accordingly, PBR thresholds are best interpreted as precautionary decision
bands rather than deterministic operational limits, particularly where
they are used to calibrate mitigation measures such as curtailment.

## Sensitivity and elasticity analysis

```{r elasticity-grid}
s_seq <- seq(0.70, 0.95, by = 0.01)
alpha_seq <- seq(1.5, 3.5, by = 0.1)

partials_lambda <- function(s, alpha, eps_s = 1e-5, eps_a = 1e-5) {
  lam <- lambda_max_niel(s, alpha)
  dlds <- (lambda_max_niel(s + eps_s, alpha) - lambda_max_niel(s - eps_s, alpha)) / (2 * eps_s)
  dlda <- (lambda_max_niel(s, alpha + eps_a) - lambda_max_niel(s, alpha - eps_a)) / (2 * eps_a)
  tibble::tibble(
    lambda_max = lam, d_lambda_ds = dlds, d_lambda_dalpha = dlda,
    E_s = dlds * s / lam, E_alpha = dlda * alpha / lam
  )
}

sens_grid <- tidyr::expand_grid(s = s_seq, alpha = alpha_seq) %>%
  rowwise() %>%
  mutate(out = list(partials_lambda(s, alpha))) %>%
  ungroup() %>%
  tidyr::unnest(out)

elasticity_ranges <- sens_grid %>%
  summarise(
    E_s_min = min(E_s), E_s_max = max(E_s), E_s_median = median(E_s),
    E_alpha_min = min(E_alpha), E_alpha_max = max(E_alpha), E_alpha_median = median(E_alpha)
  ) %>%
  tidyr::pivot_longer(everything(), names_to = c("Parameter", "Statistic"),
                       names_pattern = "E_(s|alpha)_(min|max|median)", values_to = "Value") %>%
  mutate(Parameter = recode(Parameter, s = "Adult survival (s)", alpha = "Age at first breeding (alpha)")) %>%
  tidyr::pivot_wider(names_from = Statistic, values_from = Value) %>%
  select(Parameter, min, max, median)
```

Sensitivity and elasticity analyses were undertaken to evaluate how
uncertainty in the two demographic parameters used by the Niel & Lebreton
approximation, adult survival ($s$) and age at first breeding ($\alpha$),
propagates into $\lambda_{max}$ and consequently into PBR. Parameter ranges
were set at $s \in$ [`r min(s_seq)`, `r max(s_seq)`] and $\alpha \in$
[`r min(alpha_seq)`, `r max(alpha_seq)`] years, the latter following the
literature review below. For each combination, $\lambda_{max}$ was
recalculated and local sensitivity estimated numerically using finite
differences. Elasticity is a dimensionless proportional sensitivity
metric: the expected percentage variation in $\lambda_{max}$ resulting from
a 1% variation in a demographic parameter.

```{r elasticity-table}
knitr::kable(elasticity_ranges, digits = 3, caption = "Elasticity of lambda_max to each demographic parameter.")
```

$\lambda_{max}$ is substantially more sensitive to uncertainty in adult
survival than to uncertainty in age at first breeding across the explored
demographic space: a 1% variation in adult survival produces roughly
`r round(abs(elasticity_ranges$min[elasticity_ranges$Parameter == "Adult survival (s)"]), 2)`-`r round(abs(elasticity_ranges$max[elasticity_ranges$Parameter == "Adult survival (s)"]), 2)`%
variation in $\lambda_{max}$, versus
`r round(abs(elasticity_ranges$min[elasticity_ranges$Parameter == "Age at first breeding (alpha)"]), 2)`-`r round(abs(elasticity_ranges$max[elasticity_ranges$Parameter == "Age at first breeding (alpha)"]), 2)`%
for an equivalent variation in age at first breeding. Sensitivity is not
spatially uniform: combinations of relatively high adult survival and
earlier reproductive maturity produce the largest elastic responses, which
is particularly relevant because adult survival is also among the least
well-constrained demographic parameters for most bat species.

```{r fig-elasticity-survival, fig.width=7, fig.height=4.5}
ggplot(sens_grid, aes(x = s, y = alpha, fill = E_s)) +
  geom_raster(interpolate = TRUE) +
  scale_fill_viridis_c(name = "E[s]", option = "B", direction = -1) +
  labs(x = "Adult survival (s)", y = "Age at first breeding (alpha)",
       title = "Elasticity of lambda_max to adult survival") +
  theme_minimal()
```

```{r fig-elasticity-alpha, fig.width=7, fig.height=4.5}
ggplot(sens_grid, aes(x = s, y = alpha, fill = E_alpha)) +
  geom_raster(interpolate = TRUE) +
  scale_fill_viridis_c(name = "E[alpha]", option = "A", direction = -1) +
  labs(x = "Adult survival (s)", y = "Age at first breeding (alpha)",
       title = "Elasticity of lambda_max to age at first breeding") +
  theme_minimal()
```

These results support the central rationale of the Frick adjusted
framework: uncertainty in demographic parameterisation can materially
influence resulting PBR values, and precautionary fixed-$\lambda_{max}$
assumptions reduce excessive sensitivity to uncertain demographic inputs.
However, substantial uncertainty remains even under a fixed-$\lambda_{max}$
approach, through the interaction of demographic assumptions, local
abundance estimation, and recovery-factor selection -- the remainder of
this note works through that interaction for the Parti-coloured bat at BSH
and DGY specifically.

## Evidence base for age at first breeding

For PBR, $\alpha$ should represent effective age at first reproduction --
the age at which a female actually recruits offspring into the population
-- not merely physiological sexual maturity. A literature review for
*Vespertilio murinus* was undertaken on this basis.

| Source | Species / group | Relevant information | Implication for alpha |
|---|---|---|---|
| AnAge database | *Vespertilio murinus* | Female and male sexual maturity listed at 730 days (~2 years); litter size 2; one litter per year. | Supports alpha ~= 2 as a defensible value. |
| Haensel 2010, *Nyctalus* | *Vespertilio murinus* | Some females raise young in their first year of life; the author's own data suggest most or all females may reach sexual maturity in the year of birth and raise young as one-year-olds, stated cautiously. | Suggests alpha = 1-2 may be plausible; alpha = 2 is not an optimistic assumption. |
| Ratpenats species account | *Vespertilio murinus* | Many females mate in their first autumn. | Supports early sexual maturity, but mating is not the same as confirmed breeding. |
| Racey & Entwistle synthesis, cited in Frick et al.'s own discussion | Bats broadly | The general view for most bats is that age at first breeding is around 1 year, though first-year breeding probability can be lower; Frick et al. use alpha = 1.5 as a general benchmark and suggest alpha = 2 for some families. | Supports alpha = 1.5-2 as plausible for many bat species. |
| Cryan et al. 2012 | Migratory tree bats killed at wind turbines | Ovarian development indicating sexual maturity even in first-year females for several migratory tree-roosting species. | Useful analogue: some wind-vulnerable migratory bats mature very early. |
| Komar et al. 2020 | Male *Vespertilio murinus* | Food restriction delayed seasonal sexual maturation in males; well-fed males matured earlier. | Maturation timing is condition-dependent; male data are less directly useful for PBR (which uses female recruitment). |

Table: Sources reviewed for age at first breeding in the Parti-coloured
bat.

$\alpha = 2$ is a defensible, reasonably conservative central value for PBR
purposes under the effective-recruitment reading of $\alpha$, and is
directly supported by the AnAge sexual-maturity estimate. $\alpha = 1.5$
(the paper's own general bat benchmark) is a plausible lower bound, though
it rests on thinner, more cautiously stated species-specific evidence.
$\alpha = 3-4$ has no direct empirical support for *V. murinus* in the
sources reviewed; it remains a useful precautionary sensitivity scenario
for ecological constraints on effective recruitment (migration costs, poor
juvenile survival, low first-year breeding probability), but is not the
best-supported biological estimate. This review supports using
$\alpha \in$ [1.5, 3.5] years as the working range throughout this note, in
place of a single fixed value.

## Reverse-engineering the BSH and DGY thresholds

```{r inline-threshold-values}
fr_original <- Fr_original_guess
fr_corrected <- Fr
nmin_val <- Nmin_assumed
```

BSH and DGY were each given a fixed annual fatality threshold for the
Parti-coloured bat -- 144 and 120 bats/year respectively -- without a
documented derivation. Since no independent local population estimate for
the species exists for this region, plausibility is assessed by working
backwards from the imposed thresholds under the Frick PBR equation.

```{r backsolve-table}
knitr::kable(backsolve_grid, digits = c(0, 0, 2, 0),
             col.names = c("Facility", "Threshold (bats/yr)", "Fixed lambda_max", "Required Nmin"),
             caption = "Local population size (Nmin) required to reproduce each imposed threshold, under Fr = 0.3 (Near Threatened) and the paper's two fixed lambda_max benchmarks.")
```

Pairing BSH/144 with $\lambda_{max} = 1.24$ and DGY/120 with
$\lambda_{max} = 1.20$ both back-solve to the same required
$N_{min} = `r format(nmin_val, big.mark = ",")`$ under $F_r = `r fr_original`$
(Near Threatened); no other pairing gives a shared, round value. This is
the most parsimonious explanation available for how the two thresholds
were obtained: a single assumed local population of
~`r format(nmin_val, big.mark = ",")` individuals, combined with the
paper's two $\lambda_{max}$ benchmarks applied per facility.

Whether `r format(nmin_val, big.mark = ",")` individuals is a plausible
local population for the region can be assessed via the interaction area
it implies under the paper's own default density proxy (5.5-18.2
bats/km$^2$, based on *Pipistrellus pipistrellus*, used in the absence of
species-specific density data): $N_{min} = `r format(nmin_val, big.mark = ",")`$
implies an interaction area of approximately `r round(area_for_4000)` km$^2$,
plausible as a regional buffer around two wind farms, whereas
$N_{min} = 40{,}000$ (an order of magnitude larger) would imply
approximately `r round(area_for_40000)` km$^2$ -- well beyond a
project-scale local population as the framework defines it. This does not
confirm `r format(nmin_val, big.mark = ",")` individuals as correct, but it
is the more defensible order of magnitude of the two, and is used as the
working assumption throughout the rest of this note.

$F_r = 0.3$ corresponds to a Near Threatened status. The Parti-coloured bat
is not a threatened category, so the defensible recovery factor is
$F_r = `r fr_corrected`$ (Least Concern, declining-or-unknown trend). Holding
$N_{min}$ at `r format(nmin_val, big.mark = ",")` and correcting only $F_r$:

```{r pbr-corrected-fr-table}
knitr::kable(pbr_corrected_fr, digits = 1,
             col.names = c("Fixed lambda_max", "PBR (bats/yr)"),
             caption = paste0("PBR at Nmin = ", format(nmin_val, big.mark = ","), ", Fr = ", fr_corrected, "."))
```

PBR moves from 144 to `r round(pbr_corrected_fr$PBR[pbr_corrected_fr$lambda_max_fixed == 1.24])`
($\lambda_{max} = 1.24$) and from 120 to
`r round(pbr_corrected_fr$PBR[pbr_corrected_fr$lambda_max_fixed == 1.20])`
($\lambda_{max} = 1.20$) from the recovery-factor correction alone, before
any change to $\lambda_{max}$ itself is considered.

## PBR response surfaces

```{r fig-response-surfaces, fig.width=10, fig.height=3.6}
Fr_scenarios <- c(0.1, 0.3, 0.5, 1.0)

surface_grid <- tidyr::expand_grid(
  s = seq(0.70, 0.95, length.out = 60),
  alpha = seq(1.5, 3.5, length.out = 60),
  Fr = Fr_scenarios
) %>%
  mutate(
    lambda_max = lambda_max_niel(s, alpha),
    PBR = pbr_from_components(nmin_val, Fr, lambda_max),
    Fr_label = factor(paste0("Fr = ", Fr), levels = paste0("Fr = ", Fr_scenarios))
  )

ggplot(surface_grid, aes(x = s, y = alpha, fill = PBR)) +
  geom_raster(interpolate = TRUE) +
  geom_contour(aes(z = PBR), breaks = c(120, 144), colour = "white", linewidth = 0.35) +
  facet_wrap(~Fr_label, nrow = 1) +
  scale_fill_viridis_c(name = "PBR\n(bats/yr)", option = "C") +
  labs(
    x = "Adult survival (s)", y = "Age at first breeding (alpha, yrs)",
    title = "PBR response surfaces across s, alpha and Fr",
    subtitle = paste0("Nmin = ", format(nmin_val, big.mark = ","),
                       "; white contours: PBR = 120 (DGY) and 144 (BSH)")
  ) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))
```

Across the plausible demographic range, the 120/144 contours only appear
within the $F_r = 0.3$ and $F_r = 0.5$ panels, and only for a narrow
combination of relatively high adult survival and low age at first
breeding -- outside that narrow band, PBR departs from the imposed
thresholds even before changing $F_r$.

## Monte Carlo simulation outputs

Uncertainty in adult survival and age at first breeding was propagated
into $\lambda_{max}$ and PBR by simulation, holding $N_{min} = `r format(nmin_val, big.mark = ",")`$
and $F_r = `r fr_corrected`$ fixed, with $s \sim U(0.70, 0.95)$ and
$\alpha \sim U(1.5, 3.5)$ (`r format(n_sim, big.mark = ",")` draws).

```{r mc-quantile-table}
q_tbl <- tibble::tibble(
  Quantile = names(pbr_quantiles),
  `PBR (bats/yr)` = round(pbr_quantiles, 1)
)
knitr::kable(q_tbl, caption = "Simulated PBR distribution quantiles.")
```

```{r mc-percentile-table}
knitr::kable(
  threshold_percentiles, digits = 2,
  col.names = c("Facility", "Threshold (bats/yr)", "Percentile rank in simulated PBR", "Median ratio", "P95 ratio"),
  caption = "Where the imposed thresholds fall within the simulated PBR distribution."
)
```

BSH (144) and DGY (120) sit at the
`r round(threshold_percentiles$percentile_rank_corrected[threshold_percentiles$facility == "BSH"], 0)`th
and
`r round(threshold_percentiles$percentile_rank_corrected[threshold_percentiles$facility == "DGY"], 0)`th
percentile of the simulated distribution respectively -- on the low,
conservative side, not the permissive side. Median simulated PBR is
`r round(threshold_percentiles$median_ratio[threshold_percentiles$facility == "BSH"], 2)`x
the BSH threshold and
`r round(threshold_percentiles$median_ratio[threshold_percentiles$facility == "DGY"], 2)`x
the DGY threshold; at the 95th percentile the ratios reach
`r round(threshold_percentiles$p95_ratio[threshold_percentiles$facility == "BSH"], 2)`x
and
`r round(threshold_percentiles$p95_ratio[threshold_percentiles$facility == "DGY"], 2)`x
respectively.

```{r mc-alpha-comparison-table}
knitr::kable(
  alpha_range_comparison, digits = 1,
  col.names = c("Alpha range", "Q05", "Median", "Q95", "% below 144", "% below 120"),
  caption = "Effect of the alpha range on the simulated PBR distribution (Fr = 0.5, Nmin = 4000)."
)
```

Moving from the wider, less literature-grounded $\alpha \in [2, 4]$ range
to the range the evidence review supports ($\alpha \in [1.5, 3.5]$) shifts
the simulated PBR distribution further up, not down: the literature review
therefore reinforces, rather than weakens, the conclusion below.

```{r fig-pbr-density, fig.width=7, fig.height=4}
ggplot(sim, aes(x = PBR)) +
  geom_density(fill = "grey80", colour = "grey40") +
  geom_vline(data = thresholds, aes(xintercept = threshold, colour = facility),
             linewidth = 0.8, linetype = "dashed") +
  labs(
    title = "Simulated PBR distribution vs. imposed thresholds",
    subtitle = paste0("Fr = ", fr_corrected, "; Nmin = ", format(nmin_val, big.mark = ","),
                       "; s ~ U(0.70, 0.95); alpha ~ U(1.5, 3.5)"),
    x = "PBR (bats/year)", y = "Density", colour = "Imposed threshold"
  ) +
  theme_minimal()
```

## Preliminary interpretation and implications

Under a defensible, non-optimistic parametrisation -- recovery factor
corrected to the species' actual (non-threatened) status, and demographic
parameters treated as a literature-supported range rather than a single
fixed point -- the currently imposed thresholds of 144 (BSH) and 120 (DGY)
bats/year sit near the conservative end of the resulting PBR distribution
rather than at its centre. This indicates the current thresholds are not
unreasonably permissive; if anything, a flexible parametrisation would
support materially higher values, approaching double the current figures
at the upper end of the plausible range.

These results reinforce that any operational threshold derived from PBR
should be accompanied by transparent documentation of the underlying
calculations, demographic assumptions, local population estimates, and
uncertainty analysis. A nominal threshold can still serve as a
precautionary reference point, but mitigation decisions should
additionally consider:

- the plausible uncertainty interval surrounding the threshold;
- temporal trends in fatality estimates;
- uncertainty associated with PCFM-derived mortality estimates;
- the conservation relevance and demographic sensitivity of the affected species;
- the spatial concentration and recurrence of fatalities;
- the expected effectiveness and proportionality of mitigation measures;
- and the broader adaptive-management context in which the threshold is applied.

Presenting PBR as a range rather than a single fixed number provides a
more robust and defensible basis for adaptive management decisions where
demographic parameters, local population structure, and abundance
estimates remain incompletely resolved -- as is the case for the
Parti-coloured bat in this region.
