---
title: "Bat Mortality Mitigation -- 2026 Emergency Response and 2027 Outlook (`r params$project_ref`)"
output:
  word_document:
    toc: false
params:
  project_ref: ""
  species_name: ""
  species_common_name: ""
  thresholds: !r NULL
  fr_original: !r NULL
  fr_corrected: !r NULL
  iucn_status_original_guess: ""
  iucn_status_corrected: ""
  backsolve_grid: !r NULL
  nmin_assumed: !r NULL
  project_footprint_km2: !r NULL
  combined_region_km2: !r NULL
  interproject_distance_km: !r NULL
  nmin_plausibility: !r NULL
  density_proxy_low: !r NULL
  density_proxy_high: !r NULL
  pbr_corrected_fr: !r NULL
  s_range: !r NULL
  alpha_range: !r NULL
  alpha_range_prior: !r NULL
  elasticity_ranges: !r NULL
  fig_lambda_max: ""
  fig_elasticity_survival: ""
  fig_elasticity_alpha: ""
  fig_response_surfaces: ""
  fig_response_surfaces_3d: !r NULL
  fig_response_surfaces_3d_html: !r NULL
  fig_pbr_density: ""
  n_sim: !r NULL
  pbr_quantiles: !r NULL
  threshold_percentiles: !r NULL
  alpha_range_comparison: !r NULL
  leslie_s_juv: !r NULL
  leslie_s_adult_f: !r NULL
  leslie_s_adult_m: !r NULL
  leslie_p_breed: !r NULL
  leslie_litter: !r NULL
  leslie_F: !r NULL
  lambda_leslie_validated: !r NULL
  leslie_vs_pbr: !r NULL
  fig_leslie_maturation: ""
  n0_total: !r NULL
  pva_n_years: !r NULL
  pva_n_reps: !r NULL
  pva_vital_rate_cv: !r NULL
  quasi_ext_threshold: !r NULL
  pva_risk_summary: !r NULL
  fig_pva_projection: ""
  comparison_summary: !r NULL
  leslie_breakeven_tbl: !r NULL
  leslie_boundary_band_shares: !r NULL
  leslie_boundary_pct_below_baseline: !r NULL
  leslie_boundary_range_s_adult: !r NULL
  leslie_boundary_range_s_juv: !r NULL
  leslie_boundary_range_p_breed: !r NULL
  leslie_boundary_range_litter: !r NULL
  fig_leslie_boundary_surface: ""
  nmin_vs_lambda_sensitivity: !r NULL
  fig_nmin_lambda_surface: ""
  real_correction_factor_by_project: !r NULL
  real_curtailment_start: ""
  real_curtailment_week: !r NULL
  real_checkpoint_week: !r NULL
  real_overage_summary: !r NULL
  real_yoy_summary: !r NULL
  real_forecast_summary: !r NULL
  real_n_turbines_by_project: !r NULL
  real_turbine_hist: !r NULL
  fig_real_dashboard: ""
  fig_real_turbine_pareto: ""
  turbine_validation_correction_ratio_range: !r NULL
  turbine_validation_correction_ratio_median: !r NULL
  turbine_validation_correction_ratio_by_project: !r NULL
  fig_turbine_validation: ""
  turbine_validation_bash_window: ""
  turbine_validation_dzh_window: ""
  turbine_validation_bash_data_start: ""
  turbine_validation_dzh_data_start: ""
  turbine_validation_comparison: !r NULL
  turbine_validation_overlap: !r NULL
  turbine_validation_correction_check: !r NULL
  curtailment_year_data_max_date: ""
  curtailment_year_checkpoint_week: !r NULL
  curtailment_year_reduction_so_far: !r NULL
  curtailment_year_full_counterfactual: !r NULL
  fig_curtailment_year: ""
  global_nmin_area_km2: !r NULL
  global_density_low: !r NULL
  global_density_high: !r NULL
  global_pbr_quantiles_by_project: !r NULL
  global_pbr_old_new_comparison: !r NULL
  fig_global_pbr_uncertainty: ""
  candidate_pbr_table: !r NULL
  candidate_pbr_effectiveness: !r NULL
  fig_candidate_pbr: ""
  turbine_2027_classification: !r NULL
  turbine_2027_flagged_display: !r NULL
  turbine_2027_summary_by_project: !r NULL
  turbine_2027_core_residual_summary: !r NULL
  turbine_2027_expansion_fraction: !r NULL
  turbine_2027_watchlist_n: !r NULL
  turbine_2027_candidates_project1: !r NULL
  turbine_2027_candidates_project2: !r NULL
  turbine_2027_core_residual_project1: !r NULL
  turbine_2027_core_residual_project2: !r NULL
  turbine_2027_weak_pct_project1: !r NULL
  turbine_2027_weak_pct_project2: !r NULL
  fig_turbine_2027_classification: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
options(knitr.graphics.rel_path = FALSE)
proj1 <- params$real_overage_summary$project[1]
proj2 <- params$real_overage_summary$project[2]
```

This summary draws entirely on the analysis and figures already
presented in the full technical note (*PBR Interpretation based on Frick
et al. 2026*, `r params$project_ref`); nothing here is computed
separately. It is written for a lender audience and focuses on two
questions: how effective has the 2026 emergency curtailment response
been, and what should change for 2027.

# Summary

The 2026 blanket curtailment at `r params$project_ref` was introduced as
an emergency measure after mortality had already exceeded the annual
PBR-derived reference for the year. It should therefore not be judged by
whether the 2026 annual reference was ultimately exceeded -- it was
always going to be, given how much mortality had already accumulated
before the response began. Its relevant performance period starts on
`r params$real_curtailment_start`.

Measured against the only complete untreated reference year available
(2025), mortality after implementation was substantially lower,
particularly at `r proj1`
(`r round(abs(params$curtailment_year_reduction_so_far$pct_reduction[1]))`%
below the 2025-based no-curtailment reference, versus
`r round(abs(params$curtailment_year_reduction_so_far$pct_reduction[2]))`%
at `r proj2`). This provides operational evidence that the emergency
response has restrained mortality. Because turbines were curtailed
precisely because of their high historical mortality, rather than
assigned independently of it, and only one complete untreated reference
year exists, the reduction should be read as operational effectiveness
against an explicit historical counterfactual -- not as a formal causal
estimate of curtailment's efficacy.

For 2027, the objective should shift from emergency reduction to
prospective mortality budgeting. That means: retain 2025 as the
untreated risk reference; maintain the core of turbines already shown to
carry high untreated risk; add turbines where 2026 provides recurrent,
untreated-risk evidence; and relax any existing prescription only
through a conservative, evidence-based process -- never simply because a
turbine's mortality was low while it was already being curtailed.

# 2026: two different questions

Two figures for 2026 can look inconsistent side by side unless the
underlying questions are kept separate.

**Annual mortality position.** Through
`r params$curtailment_year_data_max_date`, corrected mortality stands at
`r round(params$real_overage_summary$pct_of_threshold[1])`% of the
annual reference at `r proj1` and
`r round(params$real_overage_summary$pct_of_threshold[2])`% at `r proj2`
-- both well above 100%. This figure includes the substantial mortality
that accumulated in the months before curtailment began, when no
mitigation was in place; it describes the year's overall position
against the annual reference, not the performance of the emergency
response itself.

```{r overage-table}
knitr::kable(
  params$real_overage_summary %>%
    dplyr::select(project, threshold, corrected, pct_of_threshold),
  digits = c(0, 0, 0, 0),
  col.names = c("Project", "Annual reference", "2026 corrected mortality to date", "% of annual reference"),
  caption = paste0("2026 year-to-date mortality against the annual reference, through ", params$curtailment_year_data_max_date, ".")
)
```

**Emergency response effectiveness.** From
`r params$real_curtailment_start` onward, observed mortality is compared
against what the 2025 pattern would imply for the same weeks, had
curtailment never been introduced. This is the right basis for judging
the intervention itself.

```{r fig-curtailment-year-lender}
knitr::include_graphics(params$fig_curtailment_year)
```

```{r reduction-table}
knitr::kable(
  params$curtailment_year_reduction_so_far %>%
    dplyr::select(project, actual_corrected_total, counterfactual_corrected_total, avoided_fatalities, pct_reduction),
  digits = c(0, 0, 0, 0, 0),
  col.names = c("Project", "Observed (with curtailment)", "2025-based counterfactual (no curtailment)", "Fatalities avoided", "% reduction"),
  caption = paste0("Emergency response effectiveness, ", params$real_curtailment_start, " to ",
                    params$curtailment_year_data_max_date, ".")
)
```

Read together, these two comparisons are consistent, not contradictory:
the annual reference was already unattainable before the response began,
while the response itself, measured on its own period, shows a
substantial reduction at `r proj1` and a more moderate one at `r proj2`.

# 2027: from emergency response to prospective planning

The turbines curtailed in 2026 were chosen under a separate,
lender-agreed procedure that ranked turbines by estimated mortality over
the 2025 baseline season and selected up to 50% of that season's
cumulative mortality: 17 turbines at `r proj1`, 7 at `r proj2`. For 2027,
the same untreated 2025 evidence
remains the risk baseline, and 2026 is used only to corroborate
candidates for expansion -- never to justify removing a turbine already
curtailed, since low mortality at an already-curtailed turbine is
consistent with the intervention working, not with the turbine being
low-risk.

```{r fig-turbine-2027-lender}
knitr::include_graphics(params$fig_turbine_2027_classification)
```

```{r turbine-2027-summary-lender}
knitr::kable(
  params$turbine_2027_summary_by_project,
  digits = 0,
  col.names = c("Project", "Core (curtailed)", "Expansion candidates", "Not flagged"),
  caption = "2027 turbine classification, summary by project."
)
```

`r proj1` has two turbines (identified, not yet added) with mortality
comparable to turbines already selected in 2025, and confirmed *V.
murinus* records since curtailment began -- corroborated evidence in
both years independently, and the strongest basis available for adding
them to the 2027 prescription. `r proj2` has no turbine meeting that
bar on current evidence; this is reported as a genuine finding, not
adjusted for symmetry between the two projects. At both projects, a
minority of already-curtailed turbines still show at least one recorded
fatality since curtailment began (`r params$turbine_2027_core_residual_project1`
of 17 at `r proj1`, `r params$turbine_2027_core_residual_project2` of 7
at `r proj2`) -- evidence to consider intensifying mitigation there, not
evidence that curtailment failed.

# Recommended path forward for 2027

- Retain 2025 as the untreated risk reference; do not average or pool it
  with 2026, which would mix the reference with the effect being
  measured.
- Maintain the current core of curtailed turbines; treat any residual
  2026 fatalities among them as a signal to review intensification, not
  as grounds for relaxation.
- Add turbines where 2025 and 2026 evidence corroborate each other
  independently, as identified above.
- Relax any existing prescription only through a conservative,
  hysteresis-based process -- sustained evidence of low underlying risk
  over more than one assessment, with a predefined reinstatement
  trigger -- never on a single low-mortality season under treatment.
- Continue monitoring through the remainder of 2026 and into 2027 to
  extend the post-treatment record this assessment is based on; both the
  effectiveness comparison and the turbine classification above should
  be refreshed as further data accumulate.

This assessment does not propose any change to the imposed annual
thresholds (`r paste(sprintf("%s bats/year at %s", params$thresholds$threshold, params$thresholds$facility), collapse = "; ")`),
which remain the operative reference throughout. The full technical note
separately explores the demographic uncertainty underlying those
thresholds as background context; that discussion is not part of the
2027 operational recommendation above.
